---
title:  Structural Equation Modeling | Lab Session 3
author: A. Stefanelli^[[Alberto  Stefanelli](https://albertostefanelli.com/)], A. Alanya^[[Ahu Alanya](https://www.kuleuven.be/wieiswie/en/person/00081320)], B. Meuleman^[[Bart Meuleman](https://www.kuleuven.be/wieiswie/en/person/00041613)]
# date: "`r Sys.Date()`"
# fontsize: 10pt 
bibliography: lib.bib
link-citations: yes
output:
  #pdf_document:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    use_bookdown: true
    fig_width: 12
    fig_height: 8
# header-includes
---
<style type="text/css">
#content {
    max-width: 1500px !important;
/*    !margin-left: 300px !important;
*/
}
#table-of-contents {
    width: 300px !important;
}

#postamble {
  font-size: 10px;
}

pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}

</style>

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, 
  fig.show = 'hold', 
  fig.align = "center", 
  warning = FALSE, 
  message = FALSE, 
  comment = '')
options(width = 300, scipen = 9999)

# last dev version is needed for rmdformats.
# more info https://github.com/juba/rmdformats/issues/92 
# devtools::install_github("juba/rmdformats")

```


# What we are going to cover 

1. Measurement equivalence
2. Multi-group SEM 

# Data 

The data set used throughout is the European Social Survey ESS4-2008 Edition 4.5 was released on 1 December 2018. We will restrict the analysis to the Belgian case. Each line in the data set represents a Belgian respondent. The full dataset an documentation can be found on the [ESS website](https://www.europeansocialsurvey.org/data/download.html?r=4)

Codebook:

- **gvslvol** Standard of living for the old, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvslvue** Standard of living for the unemployed, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvhlthc** Health care for the sick, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvcldcr** Child care services for working parents, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvjbevn** Job for everyone, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvpdlwk** Paid leave from work to care for sick family, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)

- **sbstrec** Social benefits/services place too great strain on economy (1 Agree strongly - 5 Disagree strongly)
- **sbbsntx** Social benefits/services cost businesses too much in taxes/charges (1 Agree strongly - 5 Disagree strongly)
- **sbprvpv** Social benefits/services prevent widespread poverty (1 Agree strongly - 5 Disagree strongly)
- **sbeqsoc** Social benefits/services lead to a more equal society (1 Agree strongly - 5 Disagree strongly)
- **sbcwkfm** Social benefits/services make it easier to combine work and family (1 Agree strongly - 5 Disagree strongly)
- **sblazy**  Social benefits/services make people lazy (1 Agree strongly - 5 Disagree strongly)
- **sblwcoa** Social benefits/services make people less willing care for one another (1 Agree strongly - 5 Disagree strongly)
- **sblwlka** Social benefits/services make people less willing look after themselves/family (1 Agree strongly - 5 Disagree strongly)

In addition, we will use some other variables 

- **agea** Respondent's age
- **eduyrs** Years of full-time education completed
- **gndr** Gender (1 Male, 2 Female)
- **hinctnta** Household's total net income, all sources (Deciles of the actual household income range in Belgium)

- **gincdif** Government should reduce differences in income levels  (1 Agree strongly - 5 Disagree strongly)
- **dfincac** Large differences in income acceptable to reward talents and efforts  (1 Agree strongly - 5 Disagree strongly)
- **smdfslv** For fair society, differences in standard of living should be small  (1 Agree strongly - 5 Disagree strongly)


# Environment preparation 

First, let's load the necessary packages to load, manipulate, visualize and analyse the data. 

```{r, echo=T, message=FALSE, warning=FALSE,cache=F}
# Uncomment this once if you need to install the packages on your system 

### DATA MANIPULATION ###
# install.packages("haven")                 # data import from spss
# install.packages("dplyr")                 # data manipulation
# install.packages("psych")                 # descriptives
# install.packages("stringr")               # string manipulation
# install.packages("purrr")                 # table manipulation 

### MODELING ###
# install.packages("lavaan")                # SEM modelling

### VISUALIZATION ###
# install.packages("tidySEM")               # plotting SEM models

# Load the packages 

### DATA MANIPULATION ###
library("haven")        
library("dplyr")      
library("psych")
library('stringr')

### MODELING ###
library("lavaan")       

### VISUALIZATION ###
library("tidySEM")
library("purrr")  
```

# Measurament Equivalence 

- Measurement Equivalence is achieved if a measurement instrument produces equivalent results, regardless of some unrelated properties of the test subjects.
- The absence of measurement equivalence would imply some degree of distortion of the results (“bias”). For instance, an IQ test that favors males by including “gender-biased test items”
- What is the effect of such a bias? Would we draw incorrect conclusions?
- Can we detect whether such bias is present in our data?

There are different types of measurment equivalence:

1. **Configural Invariance**
    - same factor structure in each group
    - free item loadings
    - free intercepts
    - free residuals
    - No latent mean difference is estimated (fixed to 0)
2. **Metric Invariance** (also called "weak" invariance)
    - item loadings (set to equal)
    - free intercepts
    - free residuals
    - No latent mean difference is estimated (fixed to 0)
3. **Scalar Invariance** (also called "strong" invariance)
    - item loadings (set to equal)
    - intercepts (set to equal)
    - free residuals
    - latent mean difference is estimated (fixed to 0 in G1)
4. **Strict Invariance** 
    - item loadings (set to equal)
    - intercepts (set to equal)
    - residuals (fixed to 1)
    - latent mean difference is estimated (fixed to 0 in G1)
5. **Structural Invariance**
    - item loadings (set to equal)
    - intercepts (set to equal)
    - residuals (fixed to 1)
    - latent mean difference is estimated (fixed to 0 in G1)
    - factor variances (set to equal)
    - factor covariances (set to equal if the factors are more than 1)

In this lab, we are going to test whether welfare support is measured equivalently across genders. We start with configural invariance

## Configural invariance

```{r, echo=T, message=FALSE, warning=FALSE}
ess_df <- haven::read_sav("https://github.com/albertostefanelli/SEM_labs/raw/master/ESS4_belgium.sav")

# lavaan requires the grouping variable to be a factor 
# gender is coded as 1 Male, 2 Female
ess_df$gndr <- factor(ess_df$gndr,
                      levels = c("1", "2"),         # levels 
                      labels = c("Male", "Female")) # labels 

model_ws <-'welf_supp =~ gvslvol + gvslvue + gvhlthc'


fit_configural <- cfa(model_ws, 
                      data = ess_df,
                      group = "gndr")


summary(fit_configural)

```

All parameters are different across groups

## Metric Invariance

For metric invariance we are going to constrain factor loadings equal across groups. This shows that the construct has the same meaning across groups. Metric invariance is necessary for correlations and regressions. 

Metric non-invariance:

1. Meaning of the items are different across groups
2. Extreme response style might be present for some items
3. Or some people are more likely to choose a middle point for some itmes. 


```{r, echo=T, message=FALSE, warning=FALSE}
# we can simply tell lavaan what should be constrained across groups 

fit_metric <- cfa(model_ws, 
                      data = ess_df,
                      group = "gndr",
                      group.equal = c("loadings")
                )


summary(fit_metric)

```

**Q:** What (.p2.) and (.p3.) in the output refer to ?

## Scalar Invariance

Scalar invariance requires item intercepts and factor loadings equal across groups. This is important for assessing mean difference of the latent variable across groups. 

Scalar non-invariance:

1. A group tend to systematically give higher or lower item response. For instance, female respondents might express higher support for child care services. 
2. It is an additive effect. It affects the means of the observed item, hence affects the mean of the scale and the latent variable. 

```{r, echo=T, message=FALSE, warning=FALSE}

fit_scalar <- cfa(model_ws, 
                      data = ess_df,
                      group = "gndr",
                      group.equal = c("loadings",
                                      "intercepts")
                )


summary(fit_scalar)

```
## Strict invariance

For achieving strict invariance we need to constrain item residual variance,factor loadings, and intercepts equal across groups. If we want to use sum scores of observed items, our construct should reach strict invariance. This is because observed variance is a combination of true score variance and residual variance. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}

fit_strict <- cfa(model_ws, 
                      data = ess_df,
                      group = "gndr",
                      group.equal = c("intercepts",
                                      "residuals")
                )


summary(fit_strict)

```
## Structural invariance

For achieving strict invariance we need to constrain factor variances, factor covariances (if more than one latent factors), item residual variance, factor loadings, and intercepts equal across groups. Additionally, in multiple group SEM analysis we should also constrain regression path coefficients. 


```{r, echo=T, message=FALSE, warning=FALSE}

fit_structural <- cfa(model_ws, 
                      data = ess_df,
                      group = "gndr",
                      group.equal = c("loadings",
                                      "intercepts",
                                      "residuals",
                                      "lv.variances", 
                                      "lv.covariances")
                )


summary(fit_structural)

```
## Evaluating measurement invariance

As seen before there are two instruments to evaluate SEM models. Global and local fit measures.

Global Fit: 

1. Substantial decrease in goodness of fit indicates non-invariance
2. It is a good practise to look at several model fit indices rather than relying on a single one
3. Since all of the measurement invariance models are nested within one another, we can do a chi-square difference test.


```{r, echo=T, message=FALSE, warning=FALSE}
# Let's create quick function to extract the fit indices 

model_fit <-  function(lavobject) {
  vars <- c("df", "cfi", "tli", "rmsea", "rmsea.ci.lower", "rmsea.ci.upper", "rmsea.pvalue", "srmr")
  return(fitmeasures(lavobject)[vars] %>% data.frame() %>% round(2) %>% t())
}

table_fit <- 
  list(model_fit(fit_configural), 
       model_fit(fit_metric), 
       model_fit(fit_scalar), 
       model_fit(fit_strict),
       model_fit(fit_structural)) %>% 
  reduce(rbind)
 

rownames(table_fit) <- c("Configural", "Metric", "Scalar","Strict","Structural")

table_fit

# Let's compare the nested model using the anova function

table_anova <- list(anova(fit_configural, fit_metric),
       anova(fit_metric, fit_scalar),
       anova(fit_scalar, fit_strict),
       anova(fit_strict, fit_structural)) %>%  
       reduce(rbind) %>% 
       .[-c(3, 5, 7),]

table_anova
```
Remember: 

1. Invariance is achieved if there is a **non-significant** chi-square change
2. Invariance is **NOT** achieved if there is a **significant** chi-square change

**Q:** Is welfare support invariant across genders? What kind of analyses we can perform? 

## Partial invariance

Local Fit

1. The modification indices (MI) indicates the expected decrease in chi-square if a restricted parameter is to be freed in a less restrictive model
2. As done before, we can look for the largest MI value in the lavaan MI output.
3. We can free one parameter at a time through an iterative process
3. The usual cut-off value is 3.84, but this should be be adjusted based on sample size and number of tests conducted (type I error)

We have seen from the anova table that we reject the null hypothesis and that the scal invariance model fits better than the strict invariance model.  What we can further assess whether there is partial stric invariance. We can do this by using the lavTestScore() function in the lavaan function. This function allows us to see the effect of releasing equality constraints across the groups.

```{r, echo=T, message=FALSE, warning=FALSE}
lavTestScore(fit_strict)

```
The first output is a multivariate score test (i.e. Lagrange multiplier test). It indicates whether freeing all equality constraints represents an improvement in fit over the base model. In this case, we reject the null hypothesis (barely). The second part of the output is a univariate score tests (i.e., the chi-square difference tests) to see which equality constraints should be relaxed. In our case, the largest change in chi-square difference is produced by freeing the `.p5.	==	.p16` and	`.p6.	==	.p17.`. We can use the function `parTable()` to see to which parameters `.p5.	==	.p16` and	`.p6.	==	.p17.` correspond. 

```{r, echo=T, message=FALSE, warning=FALSE}
parTable(fit_strict)

```
`parTable` indicates that `.p5.	==	.p16` correspond to the the residual variance	of gvslvue, namely `gvslvue	~~	gvslvue`. As such, we start freeing the residual variance	of gvslvue.

```{r, echo=T, message=FALSE, warning=FALSE}
fit_strict_gvslvue <- cfa(model_ws, 
                      data = ess_df,
                      group = "gndr",
                      group.equal = c("loadings", 
                                      "intercepts", 
                                      "residuals"),
                      group.partial = c(gvslvue	~~	gvslvue)
                  
                )

lavTestScore(fit_strict_gvslvue)

```
Next, we run again a multivariate score test. In this case, we fail to reject multivariate score test. This means that nothing else should be freed. Let's now see if we achieved partial strict invariance by running an anova test.

```{r, echo=T, message=FALSE, warning=FALSE}

table_anova <- 
  list(anova(fit_configural, fit_metric),
       anova(fit_metric, fit_scalar),
       anova(fit_scalar, fit_strict_gvslvue)) %>%  
  reduce(rbind) %>% 
  .[-c(3, 5),]

table_anova

```

We did! 

**Q:** However, what does it means practically? Can we make comparisons based on the raw scores, if the residual/error variance is not the same across groups?

# Multi-group SEM 

Similarly to measurament invariance, we can test regression paths invariance. Multi-group SEM allows us to do it and compare the results from two or more groups. These groups could reflect experimental treatments, different locations of the data collection, different genders, or any other characteristics of interest. The goal of such an analysis is to assess whether the relationships among predictor and response variables vary by group. As such, it can be thought of as a “model-wide” interaction where we identify which paths (aka effects) change based on a group of interest (i.e. moderator). 

In this lab, we are going to fit the same model that we fit in the previous lab but in a multi-group setting

- Y is the dependent variable (Welfare support) 
- X is the predictor (Income)
- M is a mediator (Egalitarianism)
- G is a grouping variable (Gender)

First, we fit the "unconstrained" model. In this case, all the path coefficients are free to vary across groups. Since we already tested for invariance, we are going to set the loadings equal across groups.  

```{r, echo=T, message=FALSE, warning=FALSE}

model_mediation_mg <- '
## Welfare Support Factor ##
welf_supp =~ gvslvol + gvslvue + gvhlthc

## Egalitarianism ##
egual =~  gincdif + dfincac + smdfslv

## Direct effect ##
welf_supp ~ c("c1", "c2")*hinctnta

## Mediator ##
egual ~ c("a1", "a2")*hinctnta
welf_supp ~ c("b1", "b2")*egual

## Indirect effect (a*b) ##
a1b1 := a1*b1
a2b2 := a2*b2
## Total effect c + (a*b) ##
total1 := c1 + (a1*b1)
total2 := c2 + (a2*b2)

'

fit_mediation_mg <- cfa(model_mediation_mg,       # model formula
                     data = ess_df,               # data frame
                     group = "gndr",              # grouping variable (G)
                     group.equal = c("loadings")  # equal loadings 

  )

summary(fit_mediation_mg)

```

Second, we are going to fix both the loadings and the path coefficients in each group to be the same.

```{r, echo=T, message=FALSE, warning=FALSE}

model_mediation_mg_cons <- '
## Welfare Support Factor ##
welf_supp =~ gvslvol + gvslvue + gvhlthc

## Egalitarianism ##
egual =~  gincdif + dfincac + smdfslv

## Direct effect ##
welf_supp ~ c("c1", "c1")*hinctnta

## Mediator ##
egual ~ c("a1", "a1")*hinctnta
welf_supp ~ c("b1", "b1")*egual

## Indirect effect (a*b) ##
a1b1 := a1*b1
## Total effect c + (a*b) ##
total1 := c1 + (a1*b1)

'

fit_mediation_mg_cons <- cfa(model_mediation_mg_cons,       # model formula
                             data = ess_df,                 # data frame
                             group = "gndr",                # grouping variable (G)
                             group.equal = c("loadings")    # equal loadings 

  )

summary(fit_mediation_mg_cons)

```
Next, we compare the two models using the $\chi^2$ difference test:

```{r, echo=T, message=FALSE, warning=FALSE}
anova(fit_mediation_mg, fit_mediation_mg_cons)
```

The insignificant P-value implies that the unconstrained and the constrained models are not statistically significantly different. This means that the path coefficients vary very little by group. In this case, we can analyse the pooled data in a single global model.

On the contrary, a significant P-value would imply that some paths vary while others may not. If this is the case, we can introduce constraints to identify which path varies between groups. Note that the SEM model should still fit the data well, regardless of the constraints imposed on the paths. 

This exercise of relaxing and imposing constraints is potentially very exploratory. Refrain from constraining and relaxing all paths and then choosing the most parsimonious model. Instead, choosing which paths to constrain should be motivated by solid theoretical expectations and your research questions. 


