---
title:  Structural Equation Modeling | Session 1
author: D. Heerwegh, A. Alanya, A. Stefanelli, B. Meuleman
# date: "`r Sys.Date()`"
# fontsize: 10pt 
bibliography: /Users/serg/Library/Mobile Documents/com\~apple\~CloudDocs/academia/library.bib
# link-citations: yes
output:
  #pdf_document:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    use_bookdown: true
    fig_width: 12
    fig_height: 8
# header-includes
---
<style type="text/css">
#content {
    max-width: 1500px !important;
/*    !margin-left: 300px !important;
*/
}
#table-of-contents {
    width: 300px !important;
}

#postamble {
  font-size: 11px;
}

pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}

</style>

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, 
  fig.show = 'hold', 
  fig.align = "center", 
  warning = FALSE, 
  message = FALSE, 
  comment = '')
options(width = 300, scipen = 9999)

# install.packages("rmdformats")
# author: D. Heerwegh^[[Dirk Heerwegh](mailto:dirk.heerwegh@soc.kuleuven.ac.be)], A. Alanya^[[Ahu Alanya](https://www.kuleuven.be/wieiswie/en/person/00081320)], A. Stefanelli^[[Alberto  Stefanelli](https://albertostefanelli.com/)], B. Meuleman^[[B. Meuleman](https://www.kuleuven.be/wieiswie/en/person/00041613)]



```

# Lab sessions: why and how?

1. Apply theoretical knowledge
2. Increase understanding by interacting with data 3. Learn to use R and some packages within R
3. How:
    – Relatively unstructured
    – Go at your own pace, try to do the exercises yourself (do yourself a favour and do not just copy paste and run the solutions)
    - “There’s never time to do it right, but there is always time to do it over”

## Software used: **R**
    - This is not an R course!
    - We will learn some R as we go along
    - I will use RStudio
    - Many packages or libraries exist to do specific analyses

# What we are going to cover 

1. Data preparation
2. Data Exploration 
3. One-factor CFA model
    - Model fitting
    - Reviewing the output
    - Model modifications
4. A two-factor CFA model
5. Second order CFA model


# Data 

The data set used throughout is the the European Social Survey ESS4-2008 Edition 4.5 was released on 1 December 2018. We will restrict the analysis to the Belgian case.  Each line in the data set represents a Belgian respondent. The full dataset an documentation can be found on the [ESS website](https://www.europeansocialsurvey.org/data/download.html?r=4)

Codebook:

- **gvslvol** Standard of living for the old, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvslvue** Standard of living for the unemployed, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvhlthc** Health care for the sick, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvcldcr** Child care services for working parents, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvjbevn** Job for everyone, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvpdlwk** Paid leave from work to care for sick family, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)

- **sbstrec** Social benefits/services place too great strain on economy (1 Agree strongly - 5 Disagree strongly)
- **sbbsntx** Social benefits/services cost businesses too much in taxes/charges (1 Agree strongly - 5 Disagree strongly)
- **sbprvpv** Social benefits/services prevent widespread poverty (1 Agree strongly - 5 Disagree strongly)
- **sbeqsoc** Social benefits/services lead to a more equal society (1 Agree strongly - 5 Disagree strongly)
- **sbcwkfm** Social benefits/services make it easier to combine work and family (1 Agree strongly - 5 Disagree strongly)
- **sblazy**  Social benefits/services make people lazy (1 Agree strongly - 5 Disagree strongly)
- **sblwcoa** Social benefits/services make people less willing care for one another (1 Agree strongly - 5 Disagree strongly)
- **sblwlka** Social benefits/services make people less willing look after themselves/family (1 Agree strongly - 5 Disagree strongly)

- **agea** Respondent's age
- **eduyrs** Years of full-time education completed
- **gndr** Gender (1 Male, 2 Feamle)
- **hinctnta** Household's total net income, all sources (Deciles of the actual household income range in the given country.)

- **gincdif** Government should reduce differences in income levels  (1 Agree strongly - 5 Disagree strongly)
- **dfincac** Large differences in income acceptable to reward talents and efforts  (1 Agree strongly - 5 Disagree strongly)
- **smdfslv** For fair society, differences in standard of living should be small  (1 Agree strongly - 5 Disagree strongly)


# Environment preparation 

First, let's load the necessary packages to load, manipulate, visualize and analyse the data. 

```{r, echo=T, message=FALSE, warning=FALSE,cache=F}


# Uncomment this once if you need to install the packages on your system 

### DATA MANIPULATION ###
# install.packages("haven")                 # data import from spss
# install.packages("dplyr")                 # data manipulation
# install.packages("psych")                 # descriptives

# ### MODELING ###
# install.packages("lavaan")                # SEM modelling
# install.packages("lavaan.survey")         # Wrapper around packages lavaan and survey

# ### VISUALIZATION ###
# install.packages("ggplot2")               # data visualization
# install.packages("semPlot")               # plotting SEM models
# install.packages("corrplot")              # correlation plots
# install.packages("sjPlot")                # for regression output table


# Load the packages 

### DATA MANIPULATION ###
library("haven")        
library("dplyr")      
library("psych")     

### MODELING ###
library("lavaan")       
library("lavaan.survey")      

### VISUALIZATION ###
library("ggplot2")     
library("scales")  
library("corrplot")     
library("sjPlot") 

```


```{r, echo=T, message=FALSE, warning=FALSE}
ess_df <- haven::read_sav("https://github.com/albertostefanelli/SEM_labs/raw/master/ESS4_belgium.sav")

head(ess_df, 20) #first 20 rows of the dataset
nrow(ds) # number of subjects 
ncol(ds) # number of variables 
names(ds) # names of variables



```
In this first lab, we are exploring two concepts: welfare support and welfare criticism. Let's take a closer look at our items

```{r, echo=T, message=FALSE, warning=FALSE}
ess_df_selected <- ess_df %>% select(
                  # Welfare support items ##
                  gvslvol, # the old
                  gvslvue, # the unemployed
                  gvhlthc, # the sick
                  gvcldcr, # working parents
                  gvjbevn, # job for everyone
                  gvpdlwk, # paid sick leave  
                  ##	Economic criticism items ##
                  sbstrec, # strain on economy
                  sbbsntx, # too much taxes
                  ##	Social criticism items ## 
                  sbprvpv, # poverty
                  sbeqsoc, # more equal society
                  sbcwkfm, # work and family
                  ##	Moral criticism items # 
                  sblazy,  # people lazy 
                  sblwcoa, # care for others
                  sblwlka  # look after others
)

descriptive_ess <- as.data.frame(psych::describe(ess_df_selected))

descriptive_ess <- dplyr::select(descriptive_ess, 
  n,
  mean,
  sd,
  median,
  min,
  max,
  skew,
  kurtosis)

descriptive_ess

```
Before getting into more complex modelling, it is worth checking the structure of the data calculating some pairwise correlations 

```{r, echo=T, message=FALSE, warning=FALSE}

#let's first select the first 6 items concernign welfare support

ess_df_welfare_supp <- ess_df %>% select(
                  ## Welfare support items ##
                  gvslvol, # the old
                  gvslvue, # the unemployed
                  gvhlthc, # the sick
                  gvcldcr, # working parents
                  gvjbevn, # job for everyone
                  gvpdlwk  # paid sick leave  
)

# this is a correlation matrix 
welfare_supp_cor <- cor(ess_df_welfare_supp, # dataframe 
                        use = "pairwise.complete.obs" #remove NAs 
                        )

welfare_supp_cor

# visual inspection is sometimes more useful
# let's use the corrplot package 

corrplot::corrplot(welfare_supp_cor, 
         method = "circle", # magnitude of correlation as circles 
         type = "upper", # remove the bottom of the correlation matrix
         addCoef.col = "black" # add to the plot the coefficients
          )

# Discovering patterns in large correlation matrices: 
# The correlation matrix can be reordered according to the correlation coefficient. 
# This is useful for identifying the hidden structure and pattern in the matrix. 
# “hclust” for hierarchical clustering can be employed using: order="hclust"

```


```{r, echo=T, message=FALSE, warning=FALSE}

install.packages("tidySEM")
library(tidySEM)
get_layout("x", "y", rows = 1)

edg <- data.frame(from = c("f","f", "f"),
                  to = c("x1", "x2", "x3")
                  )

nod <- prepare_graph(edges = edg, layout = get_layout("", "f", "",
                                         "x1", "x2", "x3", rows = 2))


nod <- data.frame(name = c("f", "x1", "x2"),
                    shape = c("rect", "retc", "oval"),
                    label =c("A","B","C")
                    #linetype = c(2, 2, 2),
                    #colour = c("blue", "blue"),
                    #fill = c("blue", "blue"),
                    #size = c(2, 2),
                    #alpha = .5
                  )

graph_sem(edges = edg, nodes = nod, layout = get_layout("", "f", "",
                                                      "x1", "x2", "x3", rows = 2))



```

