---
title:  Structural Equation Modeling | Exercises 4
author: A. Stefanelli^[[Alberto  Stefanelli](https://albertostefanelli.com/)], A. Alanya^[[Ahu Alanya](https://www.kuleuven.be/wieiswie/en/person/00081320)], B. Meuleman^[[Bart Meuleman](https://www.kuleuven.be/wieiswie/en/person/00041613)]
# date: "`r Sys.Date()`"
# fontsize: 10pt 
bibliography: lib.bib
link-citations: yes
output:
  #pdf_document:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    use_bookdown: true
    fig_width: 12
    fig_height: 8
# header-includes
---
<style type="text/css">
#content {
    max-width: 1500px !important;
/*    !margin-left: 300px !important;
*/
}
#table-of-contents {
    width: 300px !important;
}

#postamble {
  font-size: 10px;
}

pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}

</style>

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, 
  fig.show = 'hold', 
  fig.align = "center", 
  warning = FALSE, 
  message = FALSE, 
  comment = '')
options(width = 300, scipen = 9999)

# last dev version is needed for rmdformats.
# more info https://github.com/juba/rmdformats/issues/92 
# devtools::install_github("juba/rmdformats")

```


# What we are going to cover 

1. Ex.1 -- Non-normal continuous data
2. Ex.2 -- Categorical data
3. Ex.3 -- Missing data

# Data 

The data set used throughout is the European Social Survey ESS4-2008 Edition 4.5 was released on 1 December 2018. We will restrict the analysis to the Belgian case.  Each line in the data set represents a Belgian respondent. The full dataset an documentation can be found on the [ESS website](https://www.europeansocialsurvey.org/data/download.html?r=4)

Codebook:

- **gvslvol** Standard of living for the old, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvslvue** Standard of living for the unemployed, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvhlthc** Health care for the sick, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvcldcr** Child care services for working parents, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvjbevn** Job for everyone, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvpdlwk** Paid leave from work to care for sick family, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)

- **sbstrec** Social benefits/services place too great strain on economy (1 Agree strongly - 5 Disagree strongly)
- **sbbsntx** Social benefits/services cost businesses too much in taxes/charges (1 Agree strongly - 5 Disagree strongly)
- **sbprvpv** Social benefits/services prevent widespread poverty (1 Agree strongly - 5 Disagree strongly)
- **sbeqsoc** Social benefits/services lead to a more equal society (1 Agree strongly - 5 Disagree strongly)
- **sbcwkfm** Social benefits/services make it easier to combine work and family (1 Agree strongly - 5 Disagree strongly)
- **sblazy**  Social benefits/services make people lazy (1 Agree strongly - 5 Disagree strongly)
- **sblwcoa** Social benefits/services make people less willing care for one another (1 Agree strongly - 5 Disagree strongly)
- **sblwlka** Social benefits/services make people less willing look after themselves/family (1 Agree strongly - 5 Disagree strongly)

- **agea** Respondent's age
- **eduyrs** Years of full-time education completed
- **gndr** Gender (1 Male, 2 Female)
- **hinctnta** Household's total net income, all sources (Deciles of the actual household income range in the given country.)

- **gincdif** Government should reduce differences in income levels  (1 Agree strongly - 5 Disagree strongly)
- **dfincac** Large differences in income acceptable to reward talents and efforts  (1 Agree strongly - 5 Disagree strongly)
- **smdfslv** For fair society, differences in standard of living should be small  (1 Agree strongly - 5 Disagree strongly)


# Environment preparation 

First, let's load the necessary packages to load, manipulate, visualize and analyse the data. 

```{r, echo=T, message=FALSE, warning=FALSE,cache=F}


# Uncomment this once if you need to install the packages on your system 

### DATA MANIPULATION ###
# install.packages("haven")                 # data import from spss
# install.packages("dplyr")                 # data manipulation
# install.packages("psych")                 # descriptives
# install.packages("stringr")               # string manipulation
# install.packages("purrr")                 # table manipulation 

### MODELING ###
# install.packages("lavaan")                # SEM modelling
# install.packages("MVN")                   # tests for multivariate normality
# install.packages("Amelia")                # performing multiple imputation

### VISUALIZATION ###
# install.packages("tidySEM")               # plotting SEM models
# install.packages("ggplot2")               # plotting 
# install.packages("patchwork")             # organization plots

# Load the packages 

### DATA MANIPULATION ###
library("haven")        
library("dplyr")      
library("psych")
library("stringr")
library("purrr")

### MODELING ###
library("lavaan")       
library("MVN")
library("Amelia")

### VISUALIZATION ###
library("tidySEM")
library("ggplot2")              
library("patchwork")    


```

# Ex.1 -- Non-normal continuous data

1. Check the multivariate normality for the question gincdif, dfincac, smdfslv (egalitarianism) and gvslvol gvhlthc gvcldcr gvpdlwk (welfare support) . 
2. Estimate a simple CFA (2 latent variable), without any error covariances for egalitarianism (gincdif, dfincac, smdfslv) and welfare support (gvslvol gvhlthc gvcldcr gvpdlwk)
3. Estimate the model using the ML estimation, then re-estimate using MLM. 
4. Compare fit statistics, loadings, and standard errors. 
5. Remove the covariance between egalitarianism and welfare support and re-estimate the model. Compare the fit (OPTIONAL). 

```{r, echo=T, message=FALSE, warning=FALSE}
# select the egalitarianism items
ess_df_eg <- ess_df[,c("gincdif", "dfincac", "smdfslv")]
# remove NAs
ess_df_eg_na <- na.omit(ess_df_eg)

mvn_test <- mvn(data = ess_df_eg_na, # our data without NAs
                mvnTest = c("hz")    # type of normality test to perform
                )

mvn_test$multivariateNormality


model_eg_ws <- '
## Egalitarianism ##
egual =~  gincdif + dfincac + smdfslv
## Welfare support ##
ws =~   gvslvol + gvhlthc + gvcldcr + gvpdlwk   

'

fit_eg_ws_ml <- cfa(model_eg_ws,    # model formula
                  data = ess_df,    # data frame
                  estimator = "ML"  # select the estimator 
)

summary(fit_eg_ws_ml,
         fit.measures=TRUE)

fit_eg_ws_mlr <- cfa(model_eg_ws,       # model formula
                  data = ess_df,        # data frame
                  estimator = "MLR"     # select the estimator 
)

summary(fit_eg_ws_mlr,
        fit.measures=TRUE)


tidy_results_ml <- table_results(fit_eg_ws_ml,        
  columns = c("label", 
              "est_sig", 
              "se"),
  digits = 2
)

tidy_results_mlr <- table_results(fit_eg_ws_mlr,        
  columns = c("label", 
              "est_sig", 
              "se"),
  digits = 2
)



data.frame(Parameters = tidy_results_mlr$label, 
           "ML Model" = paste(tidy_results_ml$est_sig,"(",tidy_results_ml$se, ")"),
           "MLR Model" = paste(tidy_results_mlr$est_sig,"(",tidy_results_mlr$se, ")")
)



model_eg_ws_cov <- '
## Egalitarianism ##
egual =~  gincdif + dfincac + smdfslv
## Welfare support ##
wc_socia =~ sbprvpv + sbeqsoc + sbcwkfm
ws ~~ 0*egual
'

fit_eg_ws_mlr_cov <- cfa(model_eg_ws_cov,    # model formula
                  data = ess_df,             # data frame
                  estimator = "MLR"          # select the estimator 
)


table_fit <- 
  list(model_fit(fit_eg_ws_mlr), 
       model_fit(fit_eg_ws_mlr_cov)) %>% 
  reduce(rbind)
 

rownames(table_fit) <- c("eg_ws_mlr", "eg_ws_mlr_cov")

table_fit

anova(fit_eg_ws_mlr, fit_eg_ws_mlr_cov)

```

# Ex.2 -- Measurement invariance with categorical data 

1. Re-estimate a CFA (2 latent variable) for egalitarianism (gincdif, dfincac, smdfslv) and welfare support (gvslvol gvhlthc gvcldcr gvpdlwk) treating the items as ordinal  
2. Compare the model fitted with continuous data with the one fitted with ordinal data
3. Fit a multi-group model for gender with free factor loading and corresponding threshold(s)
4. Fit a multi-group model for gender with fix factor loading and corresponding threshold(s)
5. Compare and interpret the fit. 



```{r, echo=T, message=FALSE, warning=FALSE}

model_eg_sc <- '
## Egalitarianism ##
egual =~  gincdif + dfincac + smdfslv
## Social Criticism ##
wc_socia =~ sbprvpv + sbeqsoc + sbcwkfm

'

fit_eg_sc <- cfa(model_eg_sc,        # model formula
                  data = ess_df,     # data frame
                  estimator = "WLSMV",  # select the estimator 
                  ordered = c(
                  "gincdif",
                  "dfincac",
                  "smdfslv",
                  "sbprvpv",
                  "sbeqsoc",
                  "sbcwkfm" )
)

summary(fit_eg_sc,
         fit.measures=TRUE)

fit_eg_ws_mlr <- cfa(model_eg_ws,       # model formula
                  data = ess_df,        # data frame
                  estimator = "MLR"     # select the estimator 
)

summary(fit_eg_ws_mlr,
        fit.measures=TRUE)


tidy_results_ml <- table_results(fit_eg_ws_ml,        
  columns = c("label", 
              "est_sig", 
              "se"),
  digits = 2
)

tidy_results_mlr <- table_results(fit_eg_ws_mlr,        
  columns = c("label", 
              "est_sig", 
              "se"),
  digits = 2
)



data.frame(Parameters = tidy_results_mlr$label, 
           "ML Model" = paste(tidy_results_ml$est_sig,"(",tidy_results_ml$se, ")"),
           "MLR Model" = paste(tidy_results_mlr$est_sig,"(",tidy_results_mlr$se, ")")
)



model_eg_ws_cov <- '
## Egalitarianism ##
egual =~  gincdif + dfincac + smdfslv
## Welfare support ##
ws =~   gvslvol + gvhlthc + gvcldcr + gvpdlwk   
ws ~~ 0*egual
'

fit_eg_ws_mlr_cov <- cfa(model_eg_ws_cov,    # model formula
                  data = ess_df,             # data frame
                  estimator = "MLR"          # select the estimator 
)


table_fit <- 
  list(model_fit(fit_eg_ws_mlr), 
       model_fit(fit_eg_ws_mlr_cov)) %>% 
  reduce(rbind)
 

rownames(table_fit) <- c("eg_ws_mlr", "eg_ws_mlr_cov")

table_fit

anova(fit_eg_ws_mlr, fit_eg_ws_mlr_cov)

```





Factor loadings and thresholds are fixed or freed in tandem
– If a factor loading and corresponding threshold(s) are freed, then the scale parameter needs to be
fixed to 1 • Syntax Categorical data







