---
title:  Structural Equation Modeling | Lab Session 3
author: A. Stefanelli^[[Alberto  Stefanelli](https://albertostefanelli.com/)], A. Alanya^[[Ahu Alanya](https://www.kuleuven.be/wieiswie/en/person/00081320)], B. Meuleman^[[Bart Meuleman](https://www.kuleuven.be/wieiswie/en/person/00041613)]
# date: "`r Sys.Date()`"
# fontsize: 10pt 
bibliography: lib.bib
link-citations: yes
output:
  #pdf_document:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    use_bookdown: true
    fig_width: 12
    fig_height: 8
# header-includes
---
<style type="text/css">
#content {
    max-width: 1500px !important;
/*    !margin-left: 300px !important;
*/
}
#table-of-contents {
    width: 300px !important;
}

#postamble {
  font-size: 10px;
}

pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}

</style>

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, 
  fig.show = 'hold', 
  fig.align = "center", 
  warning = FALSE, 
  message = FALSE, 
  comment = '')
options(width = 300, scipen = 9999)

# last dev version is needed for rmdformats.
# fixed these two
# more info https://github.com/juba/rmdformats/issues/92 
# more info https://github.com/juba/rmdformats/issues/93
# devtools::install_github("juba/rmdformats")

```


# What we are going to cover 

1. Data and environment preparation
2. Fit indices 
    - Local 
    - Global 
3. Ex.1 -- Local Model Fit
4. 3-factor CFA model
5. Ex.2  -- 1-factor- vs 3-factor-model 
6. Second-order model
7. Ex.3 -- Second-order model (OPTIONAL)


# Data 

The data set used throughout is the European Social Survey ESS4-2008 Edition 4.5 was released on 1 December 2018. We will restrict the analysis to the Belgian case.  Each line in the data set represents a Belgian respondent. The full dataset an documentation can be found on the [ESS website](https://www.europeansocialsurvey.org/data/download.html?r=4)

Codebook:

- **gvslvol** Standard of living for the old, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvslvue** Standard of living for the unemployed, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvhlthc** Health care for the sick, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvcldcr** Child care services for working parents, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvjbevn** Job for everyone, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvpdlwk** Paid leave from work to care for sick family, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)

- **sbstrec** Social benefits/services place too great strain on economy (1 Agree strongly - 5 Disagree strongly)
- **sbbsntx** Social benefits/services cost businesses too much in taxes/charges (1 Agree strongly - 5 Disagree strongly)
- **sbprvpv** Social benefits/services prevent widespread poverty (1 Agree strongly - 5 Disagree strongly)
- **sbeqsoc** Social benefits/services lead to a more equal society (1 Agree strongly - 5 Disagree strongly)
- **sbcwkfm** Social benefits/services make it easier to combine work and family (1 Agree strongly - 5 Disagree strongly)
- **sblazy**  Social benefits/services make people lazy (1 Agree strongly - 5 Disagree strongly)
- **sblwcoa** Social benefits/services make people less willing care for one another (1 Agree strongly - 5 Disagree strongly)
- **sblwlka** Social benefits/services make people less willing look after themselves/family (1 Agree strongly - 5 Disagree strongly)

- **agea** Respondent's age
- **eduyrs** Years of full-time education completed
- **gndr** Gender (1 Male, 2 Female)
- **hinctnta** Household's total net income, all sources (Deciles of the actual household income range in the given country.)

- **gincdif** Government should reduce differences in income levels  (1 Agree strongly - 5 Disagree strongly)
- **dfincac** Large differences in income acceptable to reward talents and efforts  (1 Agree strongly - 5 Disagree strongly)
- **smdfslv** For fair society, differences in standard of living should be small  (1 Agree strongly - 5 Disagree strongly)


# Environment preparation 

First, let's load the necessary packages to load, manipulate, visualize and analyse the data. 

```{r, echo=T, message=FALSE, warning=FALSE,cache=F}


# Uncomment this once if you need to install the packages on your system 

### DATA MANIPULATION ###
# install.packages("haven")                 # data import from spss
# install.packages("dplyr")                 # data manipulation
# install.packages("psych")                 # descriptives
# install.packages("stringr")               # string manipulation

### MODELING ###
# install.packages("lavaan")                # SEM modelling

### VISUALIZATION ###
# install.packages("tidySEM")               # plotting SEM models
# install.packages("corrplot")              # correlation/covariance plots


# Load the packages 

### DATA MANIPULATION ###
library("haven")        
library("dplyr")      
library("psych")
library('stringr')

### MODELING ###
library("lavaan")       

### VISUALIZATION ###
library("corrplot")     
library("tidySEM")

```


# Model Fit statistics 

We have two different type of fit statistics. 

1. Global fit measures:
     - They take into account how the entire entire model fit the data  
     - (some) rules of thumb: $CFI/TLI > 0.95, RMSEA < 0.05, SRMR < 0.06$  
     - current practice is: chi-square value + df + pvalue, RMSEA, CFI and SRMR
     - **DO NOT** cherry pick your fit indices
     - Check @hu_cutoff_1999
2. "Local" fit measures: 
      - looking at just one part of the model
      - usually involves the use of Modification Indexes
      - Check @thoemmes_LocalFitEvaluation_2018
    
```{r, echo=T, message=FALSE, warning=FALSE}
# import data set
ess_df <- haven::read_sav("https://github.com/albertostefanelli/SEM_labs/raw/master/ESS4_belgium.sav")

model_ws_6 <-'welf_supp =~ gvslvol + gvslvue + gvhlthc + gvcldcr + gvjbevn + gvpdlwk'


fit_ws_6 <- cfa(model_ws_6,    # model formula
                data = ess_df  # data frame
  )


## Global fit measures ##
# we can select which global fit measures to extract
fitMeasures(fit_ws_6, c("logl","AIC", "BIC", "chisq", "df", "pvalue", "cfi", "tli","rmsea"), output = "matrix")



```

**Q:** How our model perform in terms of global fit measures?

We can have a better grasp of what is happening taking a look at the local fit of our model. 

```{r, echo=T, message=FALSE, warning=FALSE}
## Local fit measures: modification indices ##
mi <- inspect(fit_ws_6,"mi")
mi.sorted <- mi[order(-mi$mi),] # sort from high to low mi.sorted[1:5,] # only display some large MI values
mi.sorted[1:5,] # only display some large MI values

# let's plot the modification indices 
plot(mi.sorted$mi) # plot the MI values
abline(h=3.84) # add a horizontal reference line (chisq value for 1 df where p=0.05)
```

**Q:** What the plot is suggesting ?

1. In certain contexts, we can modify the model based on a review of:
    - MI’s in combination with EPC’s (Expected Value Change). Both need to be “substantial”
    - Theory or the source of the data (e.g. review the content of the test items)
2. **However**, modifying a CFA moves it away from a strictly confirmatory model
    - The more modifications, the more exploratory the model becomes
    - Maybe this model was not ready for a confirmatory modelling strategy?


# Ex.1 -- Local Model Fit

1. Fit the welfare support model with 6 items
2. Modify the model by allowing error correlations and refit the model
3. Review the parameter estimates and compare to the first model
4. Review the fit statistics
5. Verify that the new model indeed fits the data better (perform a chi-squared difference test)

```{r, echo=T, message=FALSE, warning=FALSE}

model_ws_6_1 <-'welf_supp =~ gvslvol + gvslvue + gvhlthc + gvcldcr + gvjbevn + gvpdlwk'


fit_ws_6_1 <- cfa(model_ws_6_1,   # model formula
                 data = ess_df    # data frame
  )

model_ws_6_2 <-'
welf_supp =~ gvslvol + gvslvue + gvhlthc + gvcldcr + gvjbevn + gvpdlwk
gvslvol~~gvhlthc
gvcldcr~~gvpdlwk
'

fit_ws_6_2 <- cfa(model_ws_6_2,   # model formula
                  data = ess_df   # data frame
  )


fitm_model_ws_6_1 <-  fitMeasures(fit_ws_6_1, c("logl",
                                                  "AIC", 
                                                  "BIC", 
                                                  "chisq",
                                                  "df",
                                                  "pvalue",
                                                  "cfi",
                                                  "tli",
                                                  "rmsea"), output = "matrix")

fitm_model_ws_6_2 <- fitMeasures(fit_ws_6_2, c("logl",
                                                  "AIC", 
                                                  "BIC", 
                                                  "chisq",
                                                  "df",
                                                  "pvalue",
                                                  "cfi",
                                                  "tli",
                                                  "rmsea"), output = "matrix")


data.frame(Fit=rownames(fitm_model_ws_6_1),
           "model_ws_6_1" = round(fitm_model_ws_6_1[,1],2),
           "model_ws_6_2" = round(fitm_model_ws_6_2[,1],2)

  
)

anova(fit_ws_6_1,fit_ws_6_2)


```



# 3-factor CFA model 

We can also specify a model with more than 1 latent factor. Let's use the welfare criticism items. The syntax is similar to the 1-factor model. 

```{r, echo=T, message=FALSE, warning=FALSE}

model_wc <-'
## Economic criticism ##
wc_econo =~ sbstrec + sbbsntx
## Social criticism ## 
wc_socia =~ sbprvpv + sbeqsoc + sbcwkfm
##	Moral criticism ##
wc_moral =~ sblazy + sblwcoa + sblwlka
'

fit_wc <- cfa(model_wc,     # model formula
             data = ess_df  # data frame
  )

summary(fit_wc)


```


# Ex.2  -- 1-factor- vs 3-factor-model 

1. Using the welfare criticism items, fit a 1-factor model. 
2. Using the welfare criticism items, fit a 3-factor model. 
3. Verify whether a 1-factor CFA model is appropriate.
    - Review the model parameters. 
    - Compare the model fit statistics using a tabular form.
4. Modify the 3-factor model by dropping items and/or allowing error correlations and compare it with the original model (OPTIONAL)


```{r, echo=T, message=FALSE, warning=FALSE}

model_wc_1_factor <-'
wc_crit =~ sbstrec + sbbsntx + sbprvpv + sbeqsoc + sbcwkfm + sblazy + sblwcoa + sblwlka
'

fit_wc_1_factor <- cfa(model_wc_1_factor, # model formula
                      data = ess_df       # data frame
  )

summary(fit_wc_1_factor)

model_wc_3_factor <-'
## Economic criticism ##
wc_econo =~ sbstrec + sbbsntx
## Social criticism ## 
wc_socia =~ sbprvpv + sbeqsoc + sbcwkfm
##	Moral criticism ##
wc_moral =~ sblazy + sblwcoa + sblwlka
'

fit_wc_3_factor <- cfa(model_wc_3_factor,  # model formula
                      data = ess_df        # data frame
  )

summary(fit_wc_3_factor)


fitm_model_wc_1_factor <-  fitMeasures(fit_wc_1_factor, c("logl",
                                                  "AIC", 
                                                  "BIC", 
                                                  "chisq",
                                                  "df",
                                                  "pvalue",
                                                  "cfi",
                                                  "tli",
                                                  "rmsea"), output = "matrix")

fitm_model_wc_3_factor <- fitMeasures(fit_wc_3_factor, c("logl",
                                                  "AIC", 
                                                  "BIC", 
                                                  "chisq",
                                                  "df",
                                                  "pvalue",
                                                  "cfi",
                                                  "tli",
                                                  "rmsea"), output = "matrix")






data.frame("Fit Indices" = rownames(fitm_model_wc_1_factor),
           "model_wc_1_factor" = round(fitm_model_wc_1_factor[, 1], 2),
           "model_wc_3_factor" = round(fitm_model_wc_3_factor[, 1], 2)

  
)

## OPTIONAL ##

model_wc_3_factor_mod <-'
## Economic criticism ##
wc_econo =~ sbstrec + sbbsntx
## Social criticism ## 
wc_socia =~ sbprvpv + sbeqsoc + sbcwkfm
##	Moral criticism ##
wc_moral =~ sblazy + sblwcoa + sblwlka
sblwcoa ~~ sblwlka
'

fit_wc_3_factor_mod <- cfa(model_wc_3_factor_mod, # model formula
                           data = ess_df         # data frame
  )

summary(fit_wc_3_factor_mod)

fitm_model_wc_3_factor_mod <- fitMeasures(fit_wc_3_factor_mod, c("logl",
                                                  "AIC", 
                                                  "BIC", 
                                                  "chisq",
                                                  "df",
                                                  "pvalue",
                                                  "cfi",
                                                  "tli",
                                                  "rmsea"), output = "matrix")


data.frame("Fit Indices" = rownames(fitm_model_wc_1_factor),
           "model_wc_3_factor" = round(fitm_model_wc_3_factor[, 1], 2),
           "model_wc_3_factor_mod" = round(fitm_model_wc_3_factor_mod[, 1], 2)
  
)

```

# Second-order factor model 

We can also fit a second-order factor model with the three dimensions of welfare criticism.

```{r, echo=T, message=FALSE, warning=FALSE}

model_wc_2_order <-'
## Economic criticism ##
wc_econo =~ sbstrec + sbbsntx
## Social criticism ## 
wc_socia =~ sbprvpv + sbeqsoc + sbcwkfm
##	Moral criticism ##
wc_moral =~ sblazy + sblwcoa + sblwlka
## Second Order Welfare Criticism ## 
wc_crit =~ wc_econo + wc_socia + wc_moral
'

fit_wc_2_order <- cfa(model_wc_2_order,  # model formula
                      data = ess_df      # data frame
  )

```


# Ex.3 -- Second-order model (OPTIONAL)

2. Review the second-order model output using the summary function. Does the model fit well?
3. Review the $R^2$ values of the first-order latent variables. Which first-order factor is explained best by
the second-order factor?
4. Request model modification indices and explore model modifications.
5. Statistically compare model fit: does the second-order factor model fit significantly worse than the
first-order factor model?. What are the differences between the two models (e.g. degrees of freedom)?


```{r, echo=T, message=FALSE, warning=FALSE}

summary(fit_wc_2_order)

inspect(fit_wc_2_order, "r2")

mi<-inspect(fit_wc_2_order, "mi") 
# sort from high to low mi.sorted[1:5,] 
mi.sorted<-mi[order(-mi$mi), ] 
# # only display some large MI values
head(mi.sorted, 5)
```

Maybe does not makes sense to impose other modifications for this model. 

```{r, echo=T, message=FALSE, warning=FALSE}
anova(fit_wc_3_factor, fit_wc_2_order)


```

# References