---
title:  Structural Equation Modeling | Lab Session 2
author: A. Stefanelli^[[Alberto  Stefanelli](https://albertostefanelli.com/)], A. Alanya^[[Ahu Alanya](https://www.kuleuven.be/wieiswie/en/person/00081320)], B. Meuleman^[[Bart Meuleman](https://www.kuleuven.be/wieiswie/en/person/00041613)]
# date: "`r Sys.Date()`"
# fontsize: 10pt 
bibliography: lib.bib
link-citations: yes
output:
  #pdf_document:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    use_bookdown: true
    fig_width: 12
    fig_height: 8
# header-includes
---
<style type="text/css">
#content {
    max-width: 1500px !important;
/*    !margin-left: 300px !important;
*/
}
#table-of-contents {
    width: 300px !important;
}

#postamble {
  font-size: 10px;
}

pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}

</style>

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, 
  fig.show = 'hold', 
  fig.align = "center", 
  warning = FALSE, 
  message = FALSE, 
  comment = '')
options(width = 300, scipen = 9999)

# last dev version is needed for rmdformats.
# fixed these two
# more info https://github.com/juba/rmdformats/issues/92 
# more info https://github.com/juba/rmdformats/issues/93
# devtools::install_github("juba/rmdformats")

```


# What we are going to cover 

1. Data and environment preparation
2. Degrees of freedom 1-factor CFA 3 items 
3. Fit 1-factor CFA 3 items in lavaan
4. Ex 1 - One-factor CFA with 6 items
3. Initial model diagnostics and results
5. Ex 2 - Covariance matrix in lavaan
6. Plotting SEM **(OPTIONAL)**

# Data 

The data set used throughout is the European Social Survey ESS4-2008 Edition 4.5 was released on 1 December 2018. We will restrict the analysis to the Belgian case.  Each line in the data set represents a Belgian respondent. The full dataset an documentation can be found on the [ESS website](https://www.europeansocialsurvey.org/data/download.html?r=4)

Codebook:

- **gvslvol** Standard of living for the old, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvslvue** Standard of living for the unemployed, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvhlthc** Health care for the sick, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvcldcr** Child care services for working parents, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvjbevn** Job for everyone, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvpdlwk** Paid leave from work to care for sick family, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)

- **sbstrec** Social benefits/services place too great strain on economy (1 Agree strongly - 5 Disagree strongly)
- **sbbsntx** Social benefits/services cost businesses too much in taxes/charges (1 Agree strongly - 5 Disagree strongly)
- **sbprvpv** Social benefits/services prevent widespread poverty (1 Agree strongly - 5 Disagree strongly)
- **sbeqsoc** Social benefits/services lead to a more equal society (1 Agree strongly - 5 Disagree strongly)
- **sbcwkfm** Social benefits/services make it easier to combine work and family (1 Agree strongly - 5 Disagree strongly)
- **sblazy**  Social benefits/services make people lazy (1 Agree strongly - 5 Disagree strongly)
- **sblwcoa** Social benefits/services make people less willing care for one another (1 Agree strongly - 5 Disagree strongly)
- **sblwlka** Social benefits/services make people less willing look after themselves/family (1 Agree strongly - 5 Disagree strongly)

- **agea** Respondent's age
- **eduyrs** Years of full-time education completed
- **gndr** Gender (1 Male, 2 Female)
- **hinctnta** Household's total net income, all sources (Deciles of the actual household income range in the given country.)

- **gincdif** Government should reduce differences in income levels  (1 Agree strongly - 5 Disagree strongly)
- **dfincac** Large differences in income acceptable to reward talents and efforts  (1 Agree strongly - 5 Disagree strongly)
- **smdfslv** For fair society, differences in standard of living should be small  (1 Agree strongly - 5 Disagree strongly)


# Environment preparation 

First, let's load the necessary packages to load, manipulate, visualize and analyse the data. 

```{r, echo=T, message=FALSE, warning=FALSE,cache=F}


# Uncomment this once if you need to install the packages on your system 

### DATA MANIPULATION ###
# install.packages("haven")                 # data import from spss
# install.packages("dplyr")                 # data manipulation
# install.packages("psych")                 # descriptives
# install.packages("stringr")               # string manipulation

### MODELING ###
# install.packages("lavaan")                # SEM modelling

### VISUALIZATION ###
# install.packages("tidySEM")               # plotting SEM models
# install.packages("corrplot")              # correlation/covariance plots


# Load the packages 

### DATA MANIPULATION ###
library("haven")        
library("dplyr")      
library("psych")
library('stringr')

### MODELING ###
library("lavaan")       

### VISUALIZATION ###
library("corrplot")     
library("tidySEM")

```


# 1-factor CFA model 3 indicators: degrees of freedom

Now that we have an idea of what is in our data we apply our theoretical knowledge to the our empirical data. Our measurement model will be formed by 3 items that loads on the latent factor "Welfare Support".

```{r, echo=FALSE, message=FALSE, warning=FALSE}


edg <- data.frame(from = "f",
                  to = c("x1","x2","x3"),
                  label = c("l1","l2","l3")
)

nod <- data.frame(name = c("f","x1", "x2", "x3"),
                    shape = c("oval", "rect", "rect","rect"),
                    label= c("Welfare Support","gvslvol (old)","gvslvue (unemployment)"," gvhlthc (sick)"),
                    linetype = c(1,2,2,2),
                    # colour = c("blue", "blue"),
                    # fill = c("blue", "blue"),
                    # size = c(2, 2),
                     alpha = .5
                  )

graph_sem(edges = edg, nodes = nod, layout = get_layout("", "f", "",
                                                        "x1","x2","x3",
                                                        rows = 2),
           angle = 170
          )

```

Sorts of identification:

1. non-identified: $df < 0$, impossible to estimate parameters
2. over-identified: $df > 0$, we should strive for this
3. just-identified: $df = 0$, it is ok, but the fit cannot be assessed
4. empirical under-identification: it can happen when two or more indicators are highly correlated (multicollinearity)

We can quickly check if our model has positive degrees of freedom using the following formula.

**Degrees of freedom** = Number of parameters to estimate - Pieces of information

**Pieces of information** = $\dfrac{p(p+1)}{2}=\dfrac{3(3+1)}{2}=6$.

where $p$ is the number of "manifest" indicators $i$ (also called "measured variables" or "observed indicators"), in this case gvslvol gvslvue, gvhlthc. 

**Number of parameters to estimate** = $\psi{f} + \lambda_{ji} + \theta_i= 1 + 2 + 3=6$

where $\psi{f}$ is the latent factor $f$ variance, $\lambda_{fi}$ is the loading on the latent factor $f$ for each observed indicator $i$, and $\theta_{i}$ is the residual variance for each observed indicator $i$.

**Q:** Why $\lambda_{fi}=2$ ? (Pro Tip: (do not count parameters we fixed for scaling).

**Degrees of freedom** = Number of parameters to estimate - Pieces of informations =$6-6=0$

**Q:** What are the implication of having 0 degrees of freedom? 

Remember: 

- Factor variance ($\psi{j}$): how much individuals differ on the factor
- Factor loadings ($\lambda_{ji}$): the effect of the factor or latent variable on the measure
- Residual (error) variance ($\theta_{i}$) : variance in the measure not attributable to the factor

# Fitting 1-factor CFA model in Lavaan

Let's now fit our CFA model using gvslvol, gvslvue, gvhlthc. 

```{r, echo=T, message=FALSE, warning=FALSE}
ess_df <- haven::read_sav("https://github.com/albertostefanelli/SEM_labs/raw/master/ESS4_belgium.sav")

model_ws_3 <-'welf_supp =~ gvslvol + gvslvue + gvhlthc'

fit_ws_3 <- cfa(model_ws_3,    # model formula
                data = ess_df  # data frame
  )

summary(fit_ws_3)

```

## Lavaan syntax

Let's take a closer look at the lavaan syntax

- latent variable definition `=~` (is measured by)
- regression `~` (is regressed on)
- (residual) (co)variance	`~~` (is correlated with)
- intercept	`~1` (intercept)
- new model parameter `:=` (is equal to)

So, if we want, we can freely estimate the first factor loading and fix the factor variance to 1:

## Alternative model specification

Remember: Every factor in CFA should be given a metric to be identified.

1. Fix one factor loading to 1 (“marker variable” method)
2. Fix factor variance to 1
3. Fix the average loading to 1 (“effect coding” method).

```{r, echo=T, message=FALSE, warning=FALSE}

model_ws_3_alt <-'
welf_supp =~ NA*gvslvol + gvslvue + gvhlthc
welf_supp~~1*welf_supp
'

fit_ws_3_alt <- cfa(model_ws_3_alt,   # model formula
                    data = ess_df     # data frame
  )

summary(fit_ws_3_alt)

```

**Q:** What is the difference between these two models?

# Exercise 1 - 1-factor model with 6 indicators

1. Fit a model with all the 6 welfare support items 
2. Assess the degrees of freedom in this model
3. Explain the difference in degrees of freedom between the 3- and the 6-items model
3. Use the summary() function and interpret the results

```{r, echo=T, message=FALSE, warning=FALSE}
# in R 


```


# Initial model diagnostics

We can quickly check the results of our model more carefully using specific lavaan functions

1. Estimates
    - Do the indicators load well on the factor(s)?
2. Heywood cases/reasonable solution
    - Are variances positive? 
    - Are $r^2$ is below 1?
    - Are there any extreme standard errors?
3. Model fit (**next lab!**)
    - Are the fit indices good?

```{r, echo=T, message=FALSE, warning=FALSE}
# Returns the observed covariance matrix.
lavInspect(fit_ws_3, "sampstat")
# Returns the model estimated covariance matrix.
fitted(fit_ws_3)
# Returns the difference between the observed and estimated covariance matrix
residuals(fit_ws_3)
# Returns the standard errors for the free parameters in the model.
lavTech(fit_ws_3, "se")
# Returns the parameter estimates with confidence intervals.
parameterEstimates(fit_ws_3, standardized = TRUE)

```

Lavaan offers a quick way to check model statistics 

```{r, echo=T, message=FALSE, warning=FALSE}

summary(fit_ws_3,            # fitted model 
        standardized = TRUE  # indicates that we want standardized results
        )

# we can also just extract the factor loadings.
# using lavaan terminology, the factors loadings are the lambdas

inspect(fit_ws_3,           # fitted model 
        what="std")$lambda  # standardized factors loadings

# more info on the factors loading can be obtained using the tidy SEM package
# first we extract all the estimated parameters 

tidy_results <- table_results(fit_ws_3,             
  columns = c("label", 
              "est_sig", 
              "se", 
              "confint"),
  digits = 2
)

tidy_results %>% filter(str_detect(label, "welf_supp."))

# we can also take a look at residual variances 
# using lavaan terminology, the residual variances are the thetas

theta <- round(inspect(fit_ws_3, "est")$theta, 3)
theta.std <- round(inspect(fit_ws_3, "std")$theta, 3) 
r2 <- round(inspect(fit_ws_3, "r2"), 3)

data.frame(row.names = c(),                       # empty the columns names 
           Variables = colnames(theta),           # variable names 
           "Residuals" = diag(theta),             # diagonal theta
           "Std. Residuals" = diag(theta.std),    # diagonal std. theta
           "R Squared" = r2                       # R-squared
        )

```

# Ex.2 -- Calculate the model-implied covariance matrix

Using the 1-factor model with 6 indicators: 

1. Request the model-implied covariance of the matrix using lavaan’s inspect function.
2. Calculate the model-implied covariance matrix using lavaan’s inspect function.
3. Compare the two matrices. 
4. Calculate the differences between the two matrices (OPTIONAL). 

```{r, echo=T, message=FALSE, warning=FALSE}
# in R 

```


# Plotting (OPTIONAL)


```{r, echo=F, include=F, message=FALSE, warning=FALSE}
# this is just for the plot 
model_ws_6 <-'welf_supp =~ gvslvol + gvslvue + gvhlthc + gvcldcr + gvjbevn + gvpdlwk'


fit_ws_6 <- cfa(model_ws_6,    # model formula
                data = ess_df  # data frame
  )

```



Plotting SEM is quite common, especially when the model are rather complex. There are different packages that can be use to plot SEM in lavaan. We are going to use the  `tidySEM`  package. Compared to other packages, it can plot both lavaan and Mplus models and it is fairly intuitive. However, you can also check `semPlot`  or `lavaanPlot`.

```{r, echo=T, message=FALSE, warning=FALSE}
# first, let's define our plot layout 
lay <- get_layout("", "", "", "welf_supp","", "", "",
                  "", "gvslvol", "gvslvue", "gvhlthc", "gvcldcr", "gvjbevn", "gvpdlwk", rows = 2)

# let's take a look at our plot layout.
lay
# let's plot our results
plot_wc <- graph_sem(model = fit_ws_6,   # model fit
                    layout = lay,        # layout 
                    angle = 170,         # adjust the arrows 
                    label = "est_std",   # get standardized results
                    digits = 2           # round standardized results 
          )   

plot_wc
```


# References 
