---
title: Mediation Analysis with Lavaan | MENAS
author: A. Stefanelli^[[Alberto  Stefanelli](https://albertostefanelli.com/)]
# date: "`r Sys.Date()`"
# fontsize: 10pt 
bibliography: ../lib.bib
link-citations: yes
output:
  #pdf_document:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    use_bookdown: true
    fig_width: 12
    fig_height: 8
# header-includes
---
<style type="text/css">
#content {
    max-width: 1500px !important;
/*    !margin-left: 300px !important;
*/
}
#table-of-contents {
    width: 300px !important;
}

#postamble {
  font-size: 10px;
}

pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}

</style>

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, 
  fig.show = 'hold', 
  fig.align = "center", 
  warning = FALSE, 
  message = FALSE, 
  comment = '')
options(width = 300, scipen = 9999)

# last dev version is needed for rmdformats.
# more info https://github.com/juba/rmdformats/issues/92 
# devtools::install_github("juba/rmdformats")

```


# What we are going to cover 

1. MIMIC models
2. Mediation Analysis

# Data 

The data set used throughout is the European Social Survey ESS4-2008 Edition 4.5 was released on 1 December 2018. We will restrict the analysis to the Belgian case. Each line in the data set represents a Belgian respondent. The full dataset an documentation can be found on the [ESS website](https://www.europeansocialsurvey.org/data/download.html?r=4)

Codebook:

- **gvslvol** Standard of living for the old, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvslvue** Standard of living for the unemployed, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvhlthc** Health care for the sick, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvcldcr** Child care services for working parents, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvjbevn** Job for everyone, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvpdlwk** Paid leave from work to care for sick family, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)

- **sbstrec** Social benefits/services place too great strain on economy (1 Agree strongly - 5 Disagree strongly)
- **sbbsntx** Social benefits/services cost businesses too much in taxes/charges (1 Agree strongly - 5 Disagree strongly)
- **sbprvpv** Social benefits/services prevent widespread poverty (1 Agree strongly - 5 Disagree strongly)
- **sbeqsoc** Social benefits/services lead to a more equal society (1 Agree strongly - 5 Disagree strongly)
- **sbcwkfm** Social benefits/services make it easier to combine work and family (1 Agree strongly - 5 Disagree strongly)
- **sblazy**  Social benefits/services make people lazy (1 Agree strongly - 5 Disagree strongly)
- **sblwcoa** Social benefits/services make people less willing care for one another (1 Agree strongly - 5 Disagree strongly)
- **sblwlka** Social benefits/services make people less willing look after themselves/family (1 Agree strongly - 5 Disagree strongly)

In addition, we will use some other variables 

- **agea** Respondent's age
- **eduyrs** Years of full-time education completed
- **gndr** Gender (1 Male, 2 Female)
- **hinctnta** Household's total net income, all sources (Deciles of the actual household income range in Belgium)

- **gincdif** Government should reduce differences in income levels  (1 Agree strongly - 5 Disagree strongly)
- **dfincac** Large differences in income acceptable to reward talents and efforts  (1 Agree strongly - 5 Disagree strongly)
- **smdfslv** For fair society, differences in standard of living should be small  (1 Agree strongly - 5 Disagree strongly)


# Environment preparation 

First, let's load the necessary packages to load, manipulate, visualize and analyse the data. 

```{r, echo=T, message=FALSE, warning=FALSE,cache=F}
# Uncomment this once if you need to install the packages on your system 

### DATA MANIPULATION ###
# install.packages("haven")                 # data import from spss
# install.packages("dplyr")                 # data manipulation
# install.packages("psych")                 # descriptives
# install.packages("stringr")               # string manipulation

# ### MODELING ###
# install.packages("lavaan")                # SEM modelling

# ### VISUALIZATION ###
# install.packages("tidySEM")               # plotting SEM models

# Load the packages 

### DATA MANIPULATION ###
library("haven")        
library("dplyr")      
library("psych")
library('stringr')

### MODELING ###
library("lavaan")       

### VISUALIZATION ###
library("tidySEM")

```

# Data exploration

It is always a good practice to check that everything is in order and make sense of the data that we are going to analyse. 

```{r, echo=T, message=FALSE, warning=FALSE}
ess_df <- haven::read_sav("https://github.com/albertostefanelli/SEM_labs/raw/master/data/ESS4_belgium.sav")


ess_df_selected <- ess_df %>% select(
                  ## Welfare support items ##
                  gvslvol, # the old
                  gvslvue, # the unemployed
                  gvhlthc, # the sick
                  gvcldcr, # working parents
                  gvjbevn, # job for everyone
                  gvpdlwk, # paid sick leave  
                  ##	Economic criticism items ##
                  sbstrec, # strain on economy
                  sbbsntx, # too much taxes
                  ##	Moral criticism items ##
                  sblazy,  # people lazy 
                  sblwcoa, # care for others
                  sblwlka,  # look after others
                  ## Egalitarianism ##
                  gincdif,
                  dfincac,
                  smdfslv,
                  ## Demographics ##
                  agea, 
                  eduyrs,
                  gndr,
                  hinctnta


)

descriptive_ess <- as.data.frame(psych::describe(ess_df_selected))

descriptive_ess <- dplyr::select(descriptive_ess, 
  n,
  mean,
  sd,
  median,
  min,
  max,
  skew,
  kurtosis)

descriptive_ess

```

**Q:** Is everything ok ?

## Sum score indicators 

```{r, echo=T, message=FALSE, warning=FALSE}

ess_df$welf_supp <- ess_df %>% 
  dplyr::select(starts_with("gv")) %>%  
  rowwise() %>% 
  dplyr::mutate(sum = sum(cur_data(),na.rm=T)/ncol(.)) %>%  
  select(sum) %>% 
  unlist() %>% as.vector()

ess_df$welf_crit <- ess_df %>% 
  dplyr::select(starts_with("sb")) %>%  
  rowwise() %>% 
  dplyr::mutate(sum = sum(cur_data(),na.rm=T)/ncol(.)) %>%  
  select(sum) %>% 
  unlist() %>% as.vector()

ess_df$egal <- ess_df %>% 
  dplyr::select(gincdif, dfincac,smdfslv ) %>%  
  rowwise() %>% 
  dplyr::mutate(sum = sum(cur_data(),na.rm=T)/ncol(.)) %>%  
  select(sum) %>% 
  unlist() %>% as.vector()


```

# MIMIC model

Now that we constructed our indicators, we can apply our theoretical knowledge and test some simple hypotheses. We hypothesise that respondent' structural characteristics influence their support for the welfare state. These type of models are called MIMIC models and stands for “Multiple Indicators, Multiple Causes”.

```{r, echo=T, message=FALSE, warning=FALSE}

model_ws_mimic_1 <-'
welf_supp ~ welf_crit + gndr + eduyrs + hinctnta + agea
'

fit_ws_mimic_1 <- sem(model_ws_mimic_1, # model formula
                   data=ess_df 
  )

summary(fit_ws_mimic_1, 
        standardized=TRUE)


r2 <- round(inspect(fit_ws_mimic_1, "r2"),3)
r2*100

# residual variance Welfare support
sem_res_dv <- lavTech(fit_ws_mimic_1, "est", add.labels = TRUE)$psi["welf_supp","welf_supp"]
sem_res_dv
# sample-implied variance Welfare Support
sample_var <- lavTech(fit_ws_mimic_1, "sampstat", add.labels = TRUE)[[1]]$cov["welf_supp","welf_supp"]
sample_var



```

## Comparison between the ML and least square estimator

Let's compare the two different regression approaches

1. Are the coefficients the same (magnitude and significance)?
2. Is the number of degree of freedom the same across the two estimators?
3. Are we estimating the same number of parameters?
4. Are the residual variances equal?

```{r, echo=T, message=FALSE, warning=FALSE}

lm_fit_ws_mimic_1 <- lm(model_ws_mimic_1, # model formula
                       data=ess_df, 
  )

summary(lm_fit_ws_mimic_1)

```

Let's compare the estimated coefficients. 

```{r, echo=T, message=FALSE, warning=FALSE}
sem_tidy_results <- table_results(fit_ws_mimic_1,             
  columns = c("label", 
              "est_sig"),
  digits = 3,
) %>% filter(str_detect(label, "welf_supp.ON") 
                        )
sem_tidy_results$est_sig_lm <- round(coef(lm_fit_ws_mimic_1)[-1],3)

sem_tidy_results
```

The regression coefficients are identical (good!). 

One thing to note is that we don’t have an intercept in the lavaan output. This highlights an important difference between ls and ML. SEM, and more in general ML, focuses on the covariance structure of the data. However, we can include an intercept (often called mean in SEM).  

```{r, echo=T, message=FALSE, warning=FALSE}

model_ws_mimic_1 <-'
welf_supp ~ 1 + welf_crit + gndr + eduyrs + hinctnta + agea
'

fit_ws_mimic_1 <- sem(model_ws_mimic_1, # model formula
                   data=ess_df 
  )

summary(fit_ws_mimic_1, 
        standardized=TRUE)
```

Typically we include intercepts only when it’s relevant to our scientific questions. For example, when we fit multi-group models. It is worth noting that the number of free parameters increased by 1 because we were estimating an additional intercept but this does not change the degrees of freedom because we are simply adding another term to the total parameters. However, we have 7 estimated parameters and 21 piece of informations in the variance-covariance matrix so we should have a total of 8 degree of freedom instead of 0 as reported in the lavaan output. 

Let's try to understand what it is happening by looking at the model- and sample-implied covariance matrices. 

```{r, echo=T, message=FALSE, warning=FALSE}
# Returns the observed covariance matrix.
lavInspect(fit_ws_mimic_1, "sampstat")$cov
# Returns the model estimated covariance matrix.
fitted(fit_ws_mimic_1, "theta")$cov

```


There is no difference between the sample implied and the model-implied covariance matrices. This means that the fit function is equal to 0 and we are able to perfectly reproduce the structure of data. Given the fact that we have 0 degree of freedom, it is likely that we are fitting a fully saturated model where all the exogenous variables are correlated to each other. Let's check this by looking at what parameters are being estimated. 

```{r, echo=T, message=FALSE, warning=FALSE}
lavTech(fit_ws_mimic_1, "est", add.labels = TRUE)


```

Let's go through the output

1. $\Lambda$ factor loading matrix. In this case, it is empty because we are not estimating any latent factor.
2. $\Theta$ variance-covariance matrix of the residuals. Since the sample implied and model implied covariance matrices are the same, this is equal to 0. 
3. $\Phi$ model implied variance-covariance matrix. We can observe that **both variances and covariances** of our predictors are estimated!
4  $\Beta$ matrix represents the regression path between endogenous and exogenous variables. 
5. $\nu$ matrix indicates whether or not we estimate the intercept for the observed variables, 
6. $\alpha$ matrix is the sample-implied intercept matrix. 

Let's try to include the covariances among the predictors in the model syntax and check the estimated number of free parameters. 

```{r, echo=T, message=FALSE, warning=FALSE}

model_ws_mimic_2 <-'
welf_supp ~ 1 + welf_crit + gndr + eduyrs + hinctnta + agea

# covariance 
welf_crit ~~ gndr
welf_crit ~~ eduyrs
welf_crit ~~ hinctnta
welf_crit ~~ agea
gndr ~~ eduyrs
gndr ~~ hinctnta
gndr ~~ agea
eduyrs ~~ hinctnta
eduyrs ~~ agea
hinctnta ~~ agea

## res. variances 
welf_supp ~~ welf_supp
welf_crit ~~ welf_crit
gndr  ~~ gndr
eduyrs ~~ eduyrs
hinctnta ~~ hinctnta
agea ~~ agea

'

fit_ws_mimic_2 <- sem(model_ws_mimic_2, # model formula
                   data=ess_df      # data frame
  )


summary(fit_ws_mimic_2, standardized=TRUE)


```

So, we are fitting a fully saturated model where both the residual variances and the covariances are estimated. 

```{r, echo=T, message=FALSE, warning=FALSE}

semPlot::semPaths(fit_ws_mimic_1, title = FALSE, curvePivot = TRUE)

```


## Obtain the same residual variance between the LM and SEM model [OPTIONAL]

There is a discrepancy in what is reported in the `lm()` and `sem()` output concerning the residuals between. First we need to convert the residual standard error in residual variances to be allowed to compare them. 

```{r, echo=T, message=FALSE, warning=FALSE}

# model rank 
k <- length(lm_fit_ws_mimic_1$coefficients)-1 
# residual sum of square
RSS <- sum(lm_fit_ws_mimic_1$residuals**2)
# sample size 
n <- length(lm_fit_ws_mimic_1$residuals)
# residual standard error
res_se <- sqrt(RSS/(n-(1+k))) 
# residual variance
lm_res_dv <- round(res_se^2,4) 
lm_res_dv


data.frame(row.names = c(),                       # empty the columns names 
           models = c("MIMIC","lm"),
           res = c(sem_res_dv, lm_res_dv)
)
```


We can notice a discrepancy in the residuals between the two estimation procedures. Are the SEM and the LM model equivalent? Let's try to fix the covariances among the predictors to zero and see if the residual variance of Welfare Support changes. 

```{r, echo=T, message=FALSE, warning=FALSE}

model_ws_mimic_3 <-'

welf_supp ~ 1 + welf_crit + gndr + eduyrs + hinctnta + agea

# intercepts 
welf_crit ~ 0
gndr ~ 0
eduyrs ~ 0
hinctnta ~ 0
agea ~ 0


# covariance 
gndr ~~ 0*eduyrs
gndr ~~ 0*hinctnta
gndr ~~ 0*agea
eduyrs ~~ 0*hinctnta
eduyrs ~~ 0*agea
hinctnta ~~ 0*agea


'

fit_ws_mimic_3 <- sem(model_ws_mimic_3, # model formula
                   data=ess_df, # data frame
  )


summary(fit_ws_mimic_3, 
        standardized=TRUE
        )


```

The residuals are still different. This is not due to model specification but rather to how the residual variance is calculated with ML estimators. 

For OLS: $\hat{\sigma}^2_{OLS} = \frac{\sum_{i=1}^{N} \hat{\zeta_i}^2}{N-k}R$

For ML: $\hat{\sigma}^2_{ML} = \frac{\sum_{i=1}^{N} \hat{\zeta_i}^2}{N}$. 

So to make the two variance comparable: $\hat{\sigma}^2_{ML} = \left(\frac{N-k}{n}\right) \hat{\sigma}^2_{OLS}$


```{r, echo=T, message=FALSE, warning=FALSE}
sample_size <- nrow(lm_fit_ws_mimic_1$model)
# number of estimated parameters 
rank <- 6
((sample_size-rank)/sample_size)*lm_res_dv

```

The lm estimate is now equal to the one we obtain from the ML.  

*Take home:*

1. The magnitude and the significance of the fixed effects of an ordinary least squares regression is equivalent to maximum likelihood. 
2. The number of degree of freedom is calculated using the number of observations - 1 in lm. In SEM, we use the unique pieces of info in the observed covariance matrix - the number of estimated parameters. 
3. In lavaan, regression models are always fully saturated. That is, we are also estimating the covariances between all the included predictors 
4. ML and ordinary least squares estimators calculate the residual variance in different ways. 

# Mediation analysis 

In his most simple form, mediation analysis (or path analysis) tests whether the relationship between two variables is explained by a third intermediate variable. It can have a casual interpretation such as the extent to which a variable (mediator) participates in the transmittance of change from a cause to its effect. 
Consider a classical mediation setup with three variables: 

- Y is the dependent variable (Welfare support) 
- X is the predictor (Income)
- M is a mediator (Egalitarianism)

```{r, echo=FALSE, message=FALSE, warning=FALSE}


edg <- data.frame(from = c("X", "M", "X"),
                  to = c("M", "Y", "Y"),
                  label= c("a","b","c'"),
                  size = 1.5,
                  label_size = 6
)

nod <- data.frame(name = c("X","M", "Y"),
                    shape = c("rect", "rect","rect"),
                    label= c("Income","Egalitarianism","Welfare support"),
                    linetype = c(1,1,1),
                    size = c(1.5),
                    # colour = c("blue", "blue"),
                    # fill = c("blue", "blue"),
                    # size = c(2, 2),
                     alpha = .5,
                     label_size = 6
                  )

graph_sem(edges = edg, nodes = nod, layout = get_layout("", "M", "",
                                                        "X","","Y",
                                                        rows = 2),
           rect_height = 0.5,
           rect_width = 1.1,
           variance_diameter = 0.4,
           angle = 170
          )

```

This results in different paths

1. a path: Test whether X and M are significantly associated 
3. b path: Test whether M and Y are significantly associated 
1. c path: Test whether X and Y are significantly associated (**Direct Effect**)
4. c' path: Test whether Y from X are significantly associated after controlling for M (**Indirect Effect**). This is usually called "the amount of mediation".

Note that the **Total Effect** is equal to **Direct Effect** + **Indirect Effect** or $c= ab +c'$. It can be interpreted as the *pairwise correlation* between the two variables. 

Before getting into mediation analysis, let's take a closer look at the lavaan syntax for path modelling:

- regression `~` (is regressed on)
- (residual) (co)variance	`~~` (is correlated with)
- intercept	`~1` (intercept)
- new model parameter `:=` (is equal to)


```{r, echo=T, message=FALSE, warning=FALSE}

model_mediation <- '

## Direct effect ##
welf_supp ~ c*hinctnta

## Mediator ##
egal ~ a*hinctnta
welf_supp ~ b*egal

## Indirect effect (a*b) ##
ab := a*b
## Total effect ##
total := c + (a*b)

'

fit_mediation <- sem(model_mediation, # model formula
           data=ess_df                # data frame
  )

summary(fit_mediation, standardized=TRUE)

```

The indirect effect is significant and negative. We can say that egalitarianism mediates the effect between income and welfare support. However, the total effect is still not significant. 

Let's plot our model to get better grasp of what it is happening. 

```{r, echo=T, message=FALSE, warning=FALSE}
# let's organize our plot on 4 rows 
# this help our readers by having a more comprehensible plot

lay <- get_layout(
"gincdif", "dfincac", "smdfslv", "", 
"", "egal", "", "", 
"hinctnta", "", "welf_supp", "", 
"",  "gvslvol", "gvslvue", "gvhlthc",
rows = 4)

plot_mediation <- graph_sem(model = fit_mediation,   # model fit
          layout = lay,        # layout
          angle = 170          # adjust the arrows 
          #label = "est_std"   # get standardized results (not rounded)
          )

plot_mediation
```

**Q:** The path between welfare support and income (direct effect) is not significant. Nor the total effect. What does that mean ? 

**A:** We have a so called indirect-only mediation such that the indirect effect pathways fully account for the overall impact of IV on DV with the direct effect being insignificant

Zhao, Lynch and Chen [-@zhao2010reconsidering] classify mediation effects as following:

- Complementary mediation: Mediated effect (a x b) and direct effect (c) both exist and point at the same direction.
- Competitive mediation: Mediated effect (a x b) and direct effect (c) both exist and point in opposite directions.
- Indirect-only mediation: Mediated effect (a x b) exists, but no direct effect (c).
- Direct-only non-mediation: Direct effect (c) exists, but no indirect effect.
- No-effect non-mediation: Nether direct effect (c), nor indirect effect exists.

## More complex path modelling [serial mediation]

There are situation where we might want to test chain linking of the mediators, with a specified direction flow. For instance, we might hypothesise that the effect of income on Welfare Support is mediated by both Egalitarianism and Welfare Criticism. 

```{r, echo=T, message=FALSE, warning=FALSE}

model_mediation_serial <- '

## Direct effect ##
welf_supp ~ c*hinctnta
welf_supp ~ d*welf_crit

## Mediator ##
egal ~ a*hinctnta
#welf_supp ~ b*egal
welf_crit ~ e*egal

## Indirect effect ##
#ab := a*b
aed := a*e*d
## Total effect ##
total := c + (a*e*d) 



'

fit_mediation_serial <- sem(model_mediation_serial, # model formula
           data = ess_df                # data frame
  )

summary(fit_mediation_serial, standardized=TRUE)

```

Income has a negative total effect on Welfare Support ($\beta_{Total}=-0.029, -2.214, p<.05$). As theorized, this effect was serially mediated by Egalitarianism and Welfare Criticism. The indirect pathway of the effect of Income on Welfare Support via Egalitarianism and Welfare Criticism is significant and positive ($\beta_{aed}= .001, z=2.560, p<.05$). On the contrary, the direct effect of Income on Welfare support is negative ($\beta_{c}=-0.030, -2.270, p<.05$) suggesting the presence of a competitive mediation.


```{r, echo=T, message=FALSE, warning=FALSE}

semPlot::semPaths(fit_mediation_serial, 
                  title = FALSE, 
                  what= 'col', 
                  whatLabels = 'std',
                  layout= "tree2", 
                  curvePivot = TRUE, 
                  #color=c('black'),
                  edge.color=c('black'))

?semPaths

```

## Comparing nested models 

In some cases, we want to test whether adding a path between 2 predictors leads to a improvement in the fit of the model. Remember that fully saturated models have a perfect fit so it makes no sense to compare them. 

```{r, echo=T, message=FALSE, warning=FALSE}

model_mediation_serial_2 <- '

## Direct effect ##
welf_supp ~ c*hinctnta
welf_supp ~ d*welf_crit
welf_supp ~ f*egal

## Mediator ##
egal ~ a*hinctnta
#welf_supp ~ b*egal
welf_crit ~ e*egal

## Indirect effect ##
#ab := a*b
aed := a*e*d
fa := f*a
## Total effect ##
total := c + (a*e*d) 

'

fit_mediation_serial_2 <- sem(model_mediation_serial_2, # model formula
           data = ess_df                # data frame
  )

summary(fit_mediation_serial_2, 
        standardized=TRUE
        )

```


**Q:** Why we do have 1 degree of freedom ?

Let's extract the fit indices of these two model and see whether there is an improvement in the fit. 

```{r, echo=T, message=FALSE, warning=FALSE}

fitm_model_serial <-  fitMeasures(fit_mediation_serial, c("logl",
                                                "AIC", 
                                                "BIC", 
                                                "chisq",
                                                "df",
                                                "pvalue",
                                                "cfi",
                                                "srmr"), 
                                                output = "matrix")

fitm_model_serial_2 <- fitMeasures(fit_mediation_serial_2, c("logl",
                                                   "AIC", 
                                                   "BIC", 
                                                   "chisq",
                                                   "df",
                                                   "pvalue",
                                                   "cfi",
                                                   "srmr"), 
                                                   output = "matrix")



data.frame(Fit=rownames(fitm_model_serial),
           "serial" = round(fitm_model_serial[,1],2),
           "plus_f" = round(fitm_model_serial_2[,1],2)

  
)


lavaan::anova(fit_mediation_serial, fit_mediation_serial_2)

```

Adding and removing path changes the model and, consequently, the interpretation of the results. Make sure to think through this. 

## Boostrapping 

The delta method used for testing mediation is known to be problematic because the sampling distribution of the indirect path product terms is not normal. Bootstrapping is a common workaround for this problem that does not make strong assumptions about the distribution of the coefficient of interest (i.e., the sampling distributions of the two mediated paths). We can implement this using the argument `se = "bootstrap"`. 


```{r, echo=T, message=FALSE, warning=FALSE}

fit_mediation_serial_bs <- sem(model_mediation_serial_2, # model formula
           data = ess_df,
           se = "bootstrap"
  )

summary(fit_mediation_serial_bs, standardized=TRUE)

```

# !!Support Ukraine!!



# References 


