---
title:  Structural Equation Modeling | Exercise 3
author: A. Stefanelli^[[Alberto  Stefanelli](https://albertostefanelli.com/)], A. Alanya^[[Ahu Alanya](https://www.kuleuven.be/wieiswie/en/person/00081320)], B. Meuleman^[[Bart Meuleman](https://www.kuleuven.be/wieiswie/en/person/00041613)]
# date: "`r Sys.Date()`"
# fontsize: 10pt 
bibliography: lib.bib
link-citations: yes
output:
  #pdf_document:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    use_bookdown: true
    fig_width: 12
    fig_height: 8
# header-includes
---
<style type="text/css">
#content {
    max-width: 1500px !important;
/*    !margin-left: 300px !important;
*/
}
#table-of-contents {
    width: 300px !important;
}

#postamble {
  font-size: 10px;
}

pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}

</style>

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, 
  fig.show = 'hold', 
  fig.align = "center", 
  warning = FALSE, 
  message = FALSE, 
  comment = '')
options(width = 300, scipen = 9999)

# last dev version is needed for rmdformats.
# more info https://github.com/juba/rmdformats/issues/92 
# devtools::install_github("juba/rmdformats")

```


# What we are going to cover 

1. Ex.1 -- Measurement equivalence 
2. Ex.2 -- Multi-group mediation analysis 

# Data 

The data set used throughout is the European Social Survey ESS4-2008 Edition 4.5 was released on 1 December 2018. We will restrict the analysis to the Belgian case. Each line in the data set represents a Belgian respondent. The full dataset an documentation can be found on the [ESS website](https://www.europeansocialsurvey.org/data/download.html?r=4)

Codebook:

- **gvslvol** Standard of living for the old, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvslvue** Standard of living for the unemployed, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvhlthc** Health care for the sick, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvcldcr** Child care services for working parents, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvjbevn** Job for everyone, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)
- **gvpdlwk** Paid leave from work to care for sick family, governments' responsibility (0 Not governments' responsibility at all - 10 Entirely governments' responsibility)

- **sbstrec** Social benefits/services place too great strain on economy (1 Agree strongly - 5 Disagree strongly)
- **sbbsntx** Social benefits/services cost businesses too much in taxes/charges (1 Agree strongly - 5 Disagree strongly)
- **sbprvpv** Social benefits/services prevent widespread poverty (1 Agree strongly - 5 Disagree strongly)
- **sbeqsoc** Social benefits/services lead to a more equal society (1 Agree strongly - 5 Disagree strongly)
- **sbcwkfm** Social benefits/services make it easier to combine work and family (1 Agree strongly - 5 Disagree strongly)
- **sblazy**  Social benefits/services make people lazy (1 Agree strongly - 5 Disagree strongly)
- **sblwcoa** Social benefits/services make people less willing care for one another (1 Agree strongly - 5 Disagree strongly)
- **sblwlka** Social benefits/services make people less willing look after themselves/family (1 Agree strongly - 5 Disagree strongly)

In addition, we will use some other variables 

- **agea** Respondent's age
- **eduyrs** Years of full-time education completed
- **gndr** Gender (1 Male, 2 Female)
- **hinctnta** Household's total net income, all sources (Deciles of the actual household income range in Belgium)

- **gincdif** Government should reduce differences in income levels  (1 Agree strongly - 5 Disagree strongly)
- **dfincac** Large differences in income acceptable to reward talents and efforts  (1 Agree strongly - 5 Disagree strongly)
- **smdfslv** For fair society, differences in standard of living should be small  (1 Agree strongly - 5 Disagree strongly)


# Environment preparation 

First, let's load the necessary packages to load, manipulate, visualize and analyse the data. 

```{r, echo=T, message=FALSE, warning=FALSE,cache=F}
# Uncomment this once if you need to install the packages on your system 

### DATA MANIPULATION ###
# install.packages("haven")                 # data import from spss
# install.packages("dplyr")                 # data manipulation
# install.packages("psych")                 # descriptives
# install.packages("stringr")               # string manipulation
# install.packages("purrr")                 # table manipulation 

### MODELING ###
# install.packages("lavaan")                # SEM modelling

### VISUALIZATION ###
# install.packages("tidySEM")               # plotting SEM models

# Load the packages 

### DATA MANIPULATION ###
library("haven")        
library("dplyr")      
library("psych")
library('stringr')

### MODELING ###
library("lavaan")       

### VISUALIZATION ###
library("tidySEM")
library("purrr")  
```

#  Ex.1 -- Measurement equivalence 

1. Manually specify a configural invariance model for male and female
2. Manually specify a configural invariance model where you estimate residual covariances for the manifest indicators
3. Manually specify a metric invariance model for male and female
4. Manually specify a scalar invariance model for male and female
5. Explain differences in Degrees of Freedom for all the fitted models
6. Add gvcldcr and see if the model reaches scalar invariance 
7. Request modification indices for the gvcldcr scalar model and interpret them  (OPTIONAL)

Configural invariance model 

```{r, echo=T, message=FALSE, warning=FALSE}
ess_df <- haven::read_sav("https://github.com/albertostefanelli/SEM_labs/raw/master/ESS4_belgium.sav")

# lavaan requires the grouping variable to be a factor 
# gender is coded as 1 Male, 2 Female
ess_df$gndr <- factor(ess_df$gndr,
                      levels = c("1", "2"),         # levels 
                      labels = c("Male", "Female")) # labels 


model_ws_config <-'
# Factor loadings all freely estimated for each group (marker fixed to 1 by defult)
welf_supp =~ gvslvol + gvslvue + gvhlthc

# Item intercepts all freely estimated for each group
gvslvol ~ 1 
gvslvue ~ 1 
gvhlthc ~ 1 

# Residual variances all freely estimated for each group
gvslvol ~~ gvslvol 
gvslvue ~~ gvslvue 
gvhlthc ~~ gvhlthc

# Factor mean (intercept) fixed to zero in each group
welf_supp ~ 0

'


fit_configural <- cfa(model_ws_config, 
                      data = ess_df,
                      group = "gndr")


summary(fit_configural)




```
Configural invariance model with residual covariances 

```{r, echo=T, message=FALSE, warning=FALSE}

model_ws_config_cov <-'
# Factor loadings all freely estimated for each group (marker fixed to 1 by defult)
welf_supp =~ gvslvol + gvslvue + gvhlthc

# Item intercepts all freely estimated for each group
gvslvol ~ 1 
gvslvue ~ 1 
gvhlthc ~ 1 

# Residual variances all freely estimated for each group
gvslvol ~~ gvslvol 
gvslvue ~~ gvslvue 
gvhlthc ~~ gvhlthc

# Residual covariance all freely estimated for each group
gvslvol ~~ gvslvue
gvslvol ~~ gvhlthc
gvslvue ~~ gvhlthc

# Factor mean (intercept) fixed to zero in each group
welf_supp ~ 0

'


fit_configural_cov <- cfa(model_ws_config_cov, 
                      data = ess_df,
                      group = "gndr")


summary(fit_configural_cov)




```
Metric invariance model

```{r, echo=T, message=FALSE, warning=FALSE}

model_ws_metric <-'
# Factor loadings set equal
welf_supp =~ c(L1,L1)*gvslvol + c(L2,L2)*gvslvue + c(L3,L3)*gvhlthc

# Item intercepts all freely estimated for each group
gvslvol ~ 1 
gvslvue ~ 1 
gvhlthc ~ 1 

# Residual variances all freely estimated for each group
gvslvol ~~ gvslvol 
gvslvue ~~ gvslvue 
gvhlthc ~~ gvhlthc

# Factor mean (intercept) fixed to zero in each group
welf_supp ~ 0

'


fit_metric <- cfa(model_ws_metric, 
                      data = ess_df,
                      group = "gndr")


summary(fit_metric)




```
Scalar invariance model

```{r, echo=T, message=FALSE, warning=FALSE}

model_ws_scalar <-'
# Factor loadings set equal
welf_supp =~ c(L1,L1)*gvslvol + c(L2,L2)*gvslvue + c(L3,L3)*gvhlthc

# Item intercepts set equal
gvslvol ~ c(I1,I1)*1 
gvslvue ~ c(I2,I2)*1 
gvhlthc ~ c(I3,I3)*1 

# Redidual variances all freely estimated for each group
gvslvol ~~ gvslvol 
gvslvue ~~ gvslvue 
gvhlthc ~~ gvhlthc

# Factor mean (intercept) estimated in G2 and fixed to zero in G1
welf_supp ~ c(0,NA)*0

'


fit_scalar <- cfa(model_ws_scalar, 
                      data = ess_df,
                      group = "gndr")


summary(fit_scalar)


```


Scalar invariance model with the addition of gvcldcr

```{r, echo=T, message=FALSE, warning=FALSE}

model_ws_gvcldcr <- '
welf_supp =~ gvslvol + gvslvue + gvhlthc + gvcldcr
'

fit_conf_gvcldcr <- cfa(model_ws_gvcldcr, 
                      data = ess_df,
                      group = "gndr"
                )

fit_metric_gvcldcr <- cfa(model_ws_gvcldcr, 
                      data = ess_df,
                      group = "gndr",
                      group.equal = c("loadings")
                )

fit_scalar_gvcldcr <- cfa(model_ws_gvcldcr, 
                      data = ess_df,
                      group = "gndr",
                      group.equal = c("loadings",
                                      "intercepts")
                )


# Let's compare the nested model using the anova function

table_anova <- list(anova(fit_conf_gvcldcr, fit_metric_gvcldcr),
       anova(fit_metric_gvcldcr, fit_scalar_gvcldcr)) %>%  
       reduce(rbind) %>% 
       .[-c(3),]

table_anova
```
```{r, echo=T, message=FALSE, warning=FALSE}

lavTestScore(fit_scalar_gvcldcr)
parTable(fit_scalar_gvcldcr)

```

# Ex.2 -- Multi-group mediation analysis 

1. Fit a model where egalitarianism (gincdif, dfincac, smdfslv) mediates the relationship between age (agea), education (eduyrs), income (hinctnta) and welfare support (gvslvol, gvslvue, gvhlthc) in a multigroup model where all paths coefficients are free to vary across group 
    - Y is the dependent variable (Welfare support) 
    - X is a vector of predictors (age, education, income)
    - M is a mediator (Egalitarianism)
    - G is a grouping variable (Gender)
2. Assess if the model fits the data well
3. Use lavaan syntax to calculate indirect and direct effects for each predictor in each group. 
4. Interpret the difference in the path coefficients between male and female 
5. Assess if the path from Education to Egalitarianism (Path A) can be set equal across the two groups
6. Assess if the Indirect effect of Education on Welfare Support (a*b) can be set equal across the two groups (OPTIONAL)
7. Think of a method to assess if path coefficients are statistically different using the := operator (OPTIONAL)

All free 

```{r, echo=T, message=FALSE, warning=FALSE}

model_mediation_mg <- '
## Welfare Support Factor ##
welf_supp =~ gvslvol + gvslvue + gvhlthc

## Egalitarianism ##
egual =~  gincdif + dfincac + smdfslv

## Direct effect ##
welf_supp ~ c("c_inc_1", "c_inc_2")*hinctnta
welf_supp ~ c("c_age_1", "c_age_2")*agea
welf_supp ~ c("c_edu_1", "c_edu_2")*eduyrs

## Mediator ##
# Path A 
egual ~ c("a_inc_1", "a_inc_2")*hinctnta
egual ~ c("a_age_1", "a_age_2")*agea
egual ~ c("a_edu_1", "a_edu_2")*eduyrs
# Path B
welf_supp ~ c("b1", "b2")*egual

## Indirect effect (a*b) ##
# G1 
ab_inc_g1 := a_inc_1*b1
ab_age_g1 := a_age_1*b1
ab_edu_g1 := a_edu_1*b1
# G2 
ab_inc_g2 := a_inc_2*b2
ab_age_g2 := a_age_2*b2
ab_edu_g2 := a_edu_2*b2

## Total effect c + (a*b) ##
# G1 
total_inc_g1 := c_inc_1 + (a_inc_1*b1)
total_age_g1 := c_age_1 + (a_age_1*b1)
total_edu_g1 := c_edu_1 + (a_edu_1*b1)
# G1 
total_inc_g2 := c_inc_2 + (a_inc_2*b2)
total_age_g2 := c_age_2 + (a_age_2*b2)
total_edu_g2 := c_edu_2 + (a_edu_2*b2)
'

fit_mediation_mg <- cfa(model_mediation_mg,       # model formula
                     data = ess_df,               # data frame
                     group = "gndr",              # grouping variable (G)
                     group.equal = c("loadings")  # equal loadings 

  )

summary(fit_mediation_mg)

```


Path A education to Egalitarianism set equal across the two groups


```{r, echo=T, message=FALSE, warning=FALSE}

model_mediation_mg_path_a <- '
## Welfare Support Factor ##
welf_supp =~ gvslvol + gvslvue + gvhlthc

## Egalitarianism ##
egual =~  gincdif + dfincac + smdfslv

## Direct effect ##
welf_supp ~ c("c_inc_1", "c_inc_2")*hinctnta
welf_supp ~ c("c_age_1", "c_age_2")*agea
welf_supp ~ c("c_edu_1", "c_edu_2")*eduyrs

## Mediator ##
# Path A 
egual ~ c("a_inc_1", "a_inc_2")*hinctnta
egual ~ c("a_age_1", "a_age_2")*agea
egual ~ c("a_edu", "a_edu")*eduyrs
# Path B
welf_supp ~ c("b1", "b2")*egual

## Indirect effect (a*b) ##
# G1 
ab_inc_g1 := a_inc_1*b1
ab_age_g1 := a_age_1*b1
ab_edu_g1 := a_edu*b1
# G2 
ab_inc_g2 := a_inc_2*b2
ab_age_g2 := a_age_2*b2
ab_edu_g2 := a_edu*b2

## Total effect c + (a*b) ##
# G1 
total_inc_g1 := c_inc_1 + (a_inc_1*b1)
total_age_g1 := c_age_1 + (a_age_1*b1)
total_edu_g1 := c_edu_1 + (a_edu*b1)
# G1 
total_inc_g2 := c_inc_2 + (a_inc_2*b2)
total_age_g2 := c_age_2 + (a_age_2*b2)
total_edu_g2 := c_edu_2 + (a_edu*b2)
'

fit_mediation_mg_path_a <- cfa(model_mediation_mg_path_a, # model formula
                     data = ess_df,                       # data frame
                     group = "gndr",                      # grouping variable (G)
                     group.equal = c("loadings")          # equal loadings 

  )

anova(fit_mediation_mg, fit_mediation_mg_path_a)

```

Path B Egalitarianism to Welfare Support set equal across the two groups

```{r, echo=T, message=FALSE, warning=FALSE}

model_mediation_mg_path_ab <- '
## Welfare Support Factor ##
welf_supp =~ gvslvol + gvslvue + gvhlthc

## Egalitarianism ##
egual =~  gincdif + dfincac + smdfslv

## Direct effect ##
welf_supp ~ c("c_inc_1", "c_inc_2")*hinctnta
welf_supp ~ c("c_age_1", "c_age_2")*agea
welf_supp ~ c("c_edu_1", "c_edu_2")*eduyrs

## Mediator ##
# Path A 
egual ~ c("a_inc_1", "a_inc_2")*hinctnta
egual ~ c("a_age_1", "a_age_2")*agea
egual ~ c("a_edu", "a_edu")*eduyrs
# Path B
welf_supp ~ c("b", "b")*egual

## Indirect effect (a*b) ##
# G1 
ab_inc_g1 := a_inc_1*b
ab_age_g1 := a_age_1*b
ab_edu_g1 := a_edu*b
# G2 
ab_inc_g2 := a_inc_2*b
ab_age_g2 := a_age_2*b
ab_edu_g2 := a_edu*b

## Total effect c + (a*b) ##
# G1 
total_inc_g1 := c_inc_1 + (a_inc_1*b)
total_age_g1 := c_age_1 + (a_age_1*b)
total_edu_g1 := c_edu_1 + (a_edu*b)
# G1 
total_inc_g2 := c_inc_2 + (a_inc_2*b)
total_age_g2 := c_age_2 + (a_age_2*b)
total_edu_g2 := c_edu_2 + (a_edu*b)
'

fit_mediation_mg_path_ab <- cfa(model_mediation_mg_path_ab, # model formula
                     data = ess_df,                         # data frame
                     group = "gndr",                        # grouping variable (G)
                     group.equal = c("loadings")            # equal loadings 

  )

anova(fit_mediation_mg_path_a,fit_mediation_mg_path_ab)

```

Using the := for simple slope difference 

```{r, echo=T, message=FALSE, warning=FALSE}

model_mediation_mg_diff <- '
## Welfare Support Factor ##
welf_supp =~ gvslvol + gvslvue + gvhlthc

## Egalitarianism ##
egual =~  gincdif + dfincac + smdfslv

## Direct effect ##
welf_supp ~ c("c_inc_1", "c_inc_2")*hinctnta
welf_supp ~ c("c_age_1", "c_age_2")*agea
welf_supp ~ c("c_edu_1", "c_edu_2")*eduyrs

## Mediator ##
# Path A 
egual ~ c("a_inc_1", "a_inc_2")*hinctnta
egual ~ c("a_age_1", "a_age_2")*agea
egual ~ c("a_edu_1", "a_edu_2")*eduyrs
# Path B
welf_supp ~ c("b1", "b2")*egual

## Indirect effect (a*b) ##
# G1 
ab_inc_g1 := a_inc_1*b1
ab_age_g1 := a_age_1*b1
ab_edu_g1 := a_edu_1*b1
# G2 
ab_inc_g2 := a_inc_2*b2
ab_age_g2 := a_age_2*b2
ab_edu_g2 := a_edu_2*b2

## Total effect c + (a*b) ##
# G1 
total_inc_g1 := c_inc_1 + (a_inc_1*b1)
total_age_g1 := c_age_1 + (a_age_1*b1)
total_edu_g1 := c_edu_1 + (a_edu_1*b1)
# G1 
total_inc_g2 := c_inc_2 + (a_inc_2*b2)
total_age_g2 := c_age_2 + (a_age_2*b2)
total_edu_g2 := c_edu_2 + (a_edu_2*b2)
## Slope Differences ##
tot_edu_diff := total_edu_g1 - total_edu_g2

'

fit_mediation_mg_diff <- cfa(model_mediation_mg_diff, # model formula
                     data = ess_df,              # data frame
                     group = "gndr",             # grouping variable (G)
                     group.equal = c("loadings") # equal loadings 

  )

summary(fit_mediation_mg_diff)

```


